---
title: "R and RStudio"
output: github_document
---

August 30, 2018

"**The role of the teacher is to create the conditions for invention rather than provide ready-made knowledge.**"---Seymour Papert

* Today: some  basics of the R language, how to create a graph, how to put R code into the markdown document, how to generate MD and HTML from an Rmd file, problem set 1.
* Everyone should have R and RStudio on their laptops with this `02-Lesson.Rmd` notebook opened in RStudio.

## Review

### Grading

* Grades: Two exams each worth 40% and several (4-6) homework problem sets (20%).
* Exam Dates: First Exam: Tuesday, October 10, Final Exam: Friday, December 15. No makeup exams.
* Exams and problems sets will be done with R markdown documents (Rmd files) and emailed to me (both the Rmd and HTML files). 
* Exams will be done without a proctor. 
* Exams will be time restricted. 
* Exams will be available as series of questions in an Rmd file through Canvas.

### Reproducible research

If your science is to be a convincing the trail from data to final output must be openly available. R and RStudio help you generate scientific reports that are a mixture of text and code. This makes it easy to create an understandable trail from hypothesis, to experimental design, to data, to analysis, to results. Reproducible research.

### Your first graph

As motivation for learning R and RStudio, the first thing you will see is how easy it is to create a nice graph using the functions in the **ggplot2** package. Begin by obtaining the package. R has many packages. We will use several dozen of them this semester.

First download the package from a CRAN site. You can do this using the Package tab in the lower-right window pane or with the `install.packages()` function where the name of the package is in quotes inside the parentheses.
```{r eval=FALSE}
install.packages("ggplot2")
```

Next make the functions from the package available for this particular session.
```{r}
library(ggplot2)
```