---
title: "Grammar for Data"
author: "James B. Elsner"
date: September 11, 2018
output: 
  html_document:
    keep_md: true
---

"**Science is the belief in the ignorance of experts.**"---Richard Feynman

Today you will learn how to manipulate data using functions from the **dplyr** package. You will learn how to write readable code using the pipe operator. 

Data wrangling is a key component in exploratory data analysis. Exploratory data analysis preceeds data modeling.

Problem Set 3 is posted on Canvas. It's due next Tuesday (3 October).

## Revised schedule

Week | Day      | Date         | Topic
-----|----------|--------------|------
1    | Tuesday  | August 29    | Introduction
1    | Thursday | August 31    | RStudio and R
2    | Tuesday  | September 5  | Working with R
2    | Thursday | September 7  | Data and Data Frames
3    | Tuesday  | September 12 | NO CLASS IRMA
3    | Thursday | September 14 | NO CLASS IRMA
4    | Tuesday  | September 19 | NO CLASS (out of town)
4    | Thursday | September 21 | NO CLASS (out of town)
WE ARE HERE | | 
5    | Tuesday  | September 26 | Grammar for Data
5    | Thursday | September 28 | Grammar for Graphs
6    | Tuesday  | October 3    | Maps in R
6    | Thursday | October 5    | Review for First Exam
7    | Tuesday  | October 10   | FIRST EXAM
7    | Thursday | October 12   | Bayesian Data Analysis
8    | Tuesday  | October 17   | Bayesian Data Analysis
8    | Tuesday  | October 19   | Regression
9    | Tuesday  | October 24   | Regression
9    | Thursday | October 26   | Regression
10   | Tuesday  | October 31   | Classification Trees
10   | Thursday | November 2   | Logistic Regression
11   | Tuesday  | November 7   | Quantile Regression
11   | Thursday | November 9   | Grammar for Spatial Data
12   | Tuesday  | November 14  | NO CLASS (out of town)
12   | Thursday | November 16  | NO CLASS (out of town)
13   | Tuesday  | November 21  | Geographic Regression
13   | Thursday | November 23  | NO CLASS (Thanksgiving)
14   | Tuesday  | November 28  | Spatial Autocorrelation
14   | Thursday | November 30  | Spatial Regression
15   | Tuesday  | December 5   | Spatial Regression
15   | Thursday | December 7   | Review for Second Exam
16   | Friday   | December 15  | SECOND EXAM (3-5p)

### Creating a data frame

Quick review. Consider ice volume (1000 km$^3$) measurements from the arctic from 2002 to 2012. The measurements are taken on January 1st each year and are available from http://psc.apl.washington.edu/wordpress/research/projects/arctic-sea-ice-volume-anomaly/data/
```{r}
Volume = c(20.233, 19.659, 18.597, 18.948, 17.820, 
           16.736, 16.648, 17.068, 15.916, 14.455, 
           14.569)
```

The object `Volume` is a vector. Since the data values are ordered (earliest to latest) we can create another vector containing the sequence of corresponding years. Then, the function `data.frame()` creates the data frame object from the two vectors.
```{r}
Year = 2002:2012
Ice.df = data.frame(Year, Volume)
head(Ice.df)
```

The names of the columns are the names of the corresponding vectors unless otherwise specified (e.g., `yr = Year`). 

Recall we use the subset operator `[]` to extract values from the data frame. For example, what year had the minimum volume of ice?
```{r}
which.min(Ice.df$Volume)
Ice.df[10, ]
Ice.df$Year[which.min(Ice.df$Volume)]
```

To change a single vector object to a data frame use the function `as.data.frame()`. For example, let `counts` be a vector of random integers. We generate a sample of size 100 with values from a Poisson distribution having a rate of 1.69. 
```{r}
counts = rpois(100, lambda = 1.69)
head(counts)
H = as.data.frame(counts)
head(H)
```

Similarly for the ice volume data you can use the function `cbind()` [columns combine] within the function `as.data.frame()`.
```{r}
as.data.frame(cbind(Year, Volume))
```

## Grammar for data

The **dplyr** package has functions ('verbs') that perform common data operations on data frames. Filtering for rows, selecting specific columns, re-ordering rows, adding new columns, and summarizing data. The functions make working with a data frame easier than the **base** functions (`split()`, `subset()`, `lapply()`, etc).

```{r eval=FALSE}
install.packages("dplyr")
```
```{r}
library(dplyr)
```

Lets return to the `airquality` data. The data frame is in the **datasets** package and contains daily air quality measurements in New York City from May to September 1973. Ozone concentration in ppb, solar radiation in langleys, wind speed in mph, temperature in F, month and day of month.
```{r}
head(airquality)
dim(airquality)
```

There are 153 rows and 6 columns. The rows are called observations and the columns are called variables. This is typically the way you store data collected in a spreadsheet.

The notes below follow: http://genomicsclass.github.io/book/pages/dplyr_tutorial.html

The important **dplyr** verbs to remember are

Verb        | Description
-----------:|:-----------
`select()`    | selects columns; pick variables by their names
`filter()`    | filters rows; pick observations by their values
`arrange()`   | re-orders the rows
`mutate()`    | creates new columns; create new variables with functions of existing variables
`summarize()` | summarizes values; collapse many values down to a single summary
`group_by()`  | allows operations to be grouped

### Select

The function `select()` picks variables by name.

For example, create a new data frame containing only the month, day, and temperature columns from the `airquality` data frame.
```{r}
df = select(airquality, Month, Day, Temp)
head(df)
```

All verbs work the same way: The first argument is a data frame; The subsequent arguments describe what to do with the data frame, using the variable names; The result is always a new data frame

Select all the columns *except*  solar radiation.
```{r}
df = select(airquality, -Solar.R)
head(df)
```

To select a range of columns by names, use the colon (`:`) operator.
```{r}
df = select(airquality, Wind:Day)
head(df)
```

To select all columns that contain the character "n", use the function `contains()`.
```{r}
df = select(airquality, contains("n"))
head(df)
```

Additional options to select columns based on specific criteria include.

Criteria       | Description
--------------:|:-----------
`ends_with()`    | select columns that end with character string
`starts_with()`  | select columns that start with a character string
`matches()`      | select columns that match a regular expression
`one_of()`       | select column names that are from a group of names

### Filter

The function `filter()` is used to subset observations based on their values.

Create a new data frame with June values of air quality measurements.
```{r}
df = filter(airquality, Month == 6)
head(df)
```

Filter the rows for temperature at least 80 F and winds less than 5 mph.
```{r}
df = filter(airquality, Temp >= 80 & Wind < 5)
head(df)
```

Multiple arguments to `filter()` are combined with "and": every expression must be true in order for a row to be included in the output. 

For other types of combinations, we need to use Boolean operators: `&` is "and", `|` is "or", and `!` is "not".

The following code finds all observations taken in May and September.
```{r}
df = filter(airquality, Month == 5 | Month == 9)
```

Note: You can't write `filter(airquality, Month == 5 | 9)` although `filter(airquality, Month %in% c(5, 9))` would work. 

We can use the operators (e.g., `>`, `<`, `>=`, `<=`, `!=`, `|` (or)) to create logical tests for filtering.

The easiest mistake to make is to use `=` instead of `==` when testing for equality.

### The pipe operator

Above, with each function you assign the result to a new data frame and then you print the first six rows of the new data frame with the function `head()`. You can nest function like this
```{r}
head(filter(airquality, Temp >= 80 & Wind < 5))
```

This is a bit hard to read as you are reading from inside out. The pipe operator `%>%` from the **magrittr** package (installed with **dplyr**) allows you to direct (pipe) the output from one function to the input of the next function. It produces code that is easier to read.

For the example above, you start with the name of the data frame followed by the pipe operator followed by the action verb. You read this as "Take the data frame `airquality` THEN filter for `Temp` at least 80 AND `Wind` less than 5, THEN list the first six rows."
```{r}
airquality %>%
  filter(Temp >= 80 & Wind < 5) %>%
  head()
```

The output from the function (or object) to the left of the pipe operator is assumed as the first argument of the function to the right of the pipe. It involves a bit more code but it's simpler to read. 

Another example
```{r}
airquality %>% 
  select(Temp, Day, Month) %>% 
  head()
```

"Take the data frame `airquality` THEN select the columns `Temp`, `Day`, and `Month`, THEN list the first six rows."

This is handy when you want to create a new data frame that is a filtered, subsetted, etc version of the original one. For example, suppose you only want to work with the ozone, wind and temperature data from August and September when the winds are light.
```{r}
df = airquality %>%
  filter(Month == 8 | Month == 9 & Wind < 5) %>%
  select(Month, Day, Ozone, Wind, Temp)
```

"Assign to `df` the `airquality` data frame filtered by month and wind speed and selecting only the columns labeled `Month`, `Day`, `Ozone`, `Wind`, and `Temp`"
  
### Arrange

To arrange (or re-order) rows by a particular column such as solar radiation.
```{r}
airquality %>%
  arrange(desc(Solar.R)) %>%
  head()
```

Now select three columns from `airquality`, arrange the rows by temperature and then arrange by solar radiation. Assign this manipulation to a new data frame.
```{r}
df = airquality %>%
  select(Solar.R, Wind, Temp) %>%
  arrange(Temp, Solar.R)
head(df)
```

The arrangement is done from lowest to highest values first by temperature and then by radiation.

Same as above, except here you filter for temperatures exceeding 90 F.
```{r}
airquality %>%
  select(Solar.R, Wind, Temp) %>%
  arrange(Temp, Solar.R) %>%
  filter(Temp > 90)
```

Same as above, except here you arrange by descending temperature with the `desc()` function.
```{r}
airquality %>%
  select(Solar.R, Wind, Temp) %>%
  arrange(desc(Temp), Solar.R) %>%
  filter(Temp > 90)
```

The second variable breaks the ties from the first variable.

### Mutate

The `mutate()` function adds new columns to the data frame. For example, create a new column called `TempC` which is the temperature in Celcius.
```{r}
airquality %>%
  mutate(TempC = (Temp - 32) * 5/9,
         WindMS = Wind * .44704) %>%
  head()
```

On days when the temperature is below 60 F compute the apparent temperature based on wind chill and arrange from coldest to warmest apparent temperature.
```{r}
airquality %>%
  filter(Temp < 60) %>%
  mutate(TempAp = 35.74 + .6215 * Temp - 35.75 * Wind^.16 + .4275 * Temp * Wind^.16) %>%
  arrange(TempAp)
```

### Summarize

The `summarize()` (or `summarise()`) function creates summary statistics for a given column in the data frame. To compute the average wind speed during July type
```{r}
airquality %>%
  filter(Month == 7) %>%
  summarize(Wavg = mean(Wind))
```

We've looked at several summary functions already including `sum()`, `sd()`, `min()`, `max()`, `var()`, `range()`, `median()`. Others include

Summary function  | Description
-----------------:|:-----------
`n()`               | Length of the column
`first()`           | First value of the column
`last()`           | Last value of the column
`n_distinct()`      | Number of distinct values

Find the maximum and median wind speed and maximum ozone concentration values during May. Also determine the number of days.
```{r}
airquality %>%
  filter(Month == 5) %>%
  summarize(Wmax = max(Wind),
            Wmed = median(Wind),
            OzoneMax = max(Ozone, na.rm = TRUE),
            NumDays = n())
```

### Grouping

What if you want to do the above summary separately for each month? Use the `group_by()` function. You split the data frame by some variable (e.g., `Month`), apply a function to the individual data frames, and then combine the output.

Find the highest ozone concentration for each month.
```{r}
airquality %>%
  group_by(Month) %>%
  summarize(OzoneMax =  max(Ozone, na.rm = TRUE),
            NumDays = n())
```

Find the average ozone concentration when temperatures are above and below 70 F.
```{r}
airquality %>%
  group_by(Temp >= 70) %>%
  summarize(OzoneAvg =  mean(Ozone, na.rm = TRUE),
            NumDays = n())
```

On average ozone concentrations are higher on warm days (Temp >= 70 F) days. That is to say mean ozone concentration depends on temperature, where the word 'depends' implies statistical dependency.

The mean is a model for the data. The statistical dependency of the mean implies that a model for ozone concentration will likely be improved by including temperature as an explanatory variable.

### A grammar

As noted the syntax of the verb functions are all the same:

* The first argument is a data frame.
* The subsequent arguments describe what to do with the data frame. Notice that you can refer to columns in the data frame directly without using $.
* The result is a new data frame

These properties make it easy to chain together many simple steps to achieve a complex result.

The five functions provide the basis of a grammar for data. At the most basic level, you can only alter a data frame in five useful ways: you can reorder the rows (`arrange()`), pick observations and variables of interest (`filter()` and `select()`), add new variables that are functions of existing variables (`mutate()`), or collapse many values to a summary (`summarise()`).

## New York City flight data

Let's apply these operations to a larger set of data. The data frame is in the package **nycflights13**. The data set is called `flights`. The notes follow closely from https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html
```{r}
library(nycflights13)
dim(flights)
```

The data contains all 336,776 flights that departed NYC in 2013 and comes from the U.S. Bureau of Transporation Statistics. More information is available by typing `?nycflights13`.

The object flights is actually a tibble (tabled data frame). When you are dealing with a large data frame it is useful to make it a tibble so that you don't accidently print it to the screen.

For example, airquality is not a tibble. So when you type the name of the data frame the entire data frame scrolls in the console. To change a data frame to a tibble, use the function `tbl_df()`.
```{r}
airquality
airquality = tbl_df(airquality)
airquality
```

Returning to the NYC flights data.
```{r}
flights
```

The function `filter()` allows you to select a subset of rows in a data frame 

How would you extract all flights occurring on February 1st?
```{r}
flights %>%
  filter(month == 2 & day == 1)
```

The function `arrange()` works like `filter()` except that it reorders the rows. If you provide more than one column name, each additional column will be used to break ties in the values of the preceding columns.

How would you arrange all flights in descending order of departure delay?
```{r}
flights %>%
  arrange(desc(dep_delay))
```

Often you work with large datasets with many columns but only a few may be of interest. The function `select()` allows you to zoom in on an interesting subset of the columns.

How would you create a data frame containing only the dates, carrier, and flight numbers?
```{r}
df = flights %>%
  select(year:day, carrier, flight)
df
```

You add new columns with the function `mutate()`.

Compute the time gained in flight by subtracting the departure delay (minutes) from the arrival delay.
```{r}
flights %>%
  mutate(gain = arr_delay - dep_delay) %>%
  select(year:day, carrier, flight, gain) %>%
  arrange(desc(gain))
```

Determine the mean departure delay.
```{r}
flights %>%
  summarize(avgDelay = mean(dep_delay, na.rm = TRUE))
```

### Randomly sample rows

You use `sample_n()` and `sample_frac()` to take random sample of rows from your data frame. Take a random sample of five rows from the flights data frame.
```{r}
flights %>%
  sample_n(5)
```

Take a random sample of 1% of the rows.
```{r}
flights %>%
  sample_frac(.01)
```

Use the argument `replace = TRUE` to perform a bootstrap sample. More on this later.

The verbs are powerful when you apply them to groups of observations within a data frame. This is done with the function `group_by()`. Determine the average arrival delay by airplane (tail number).
```{r}
flights %>%
  group_by(tailnum) %>%
  summarize(delayAvg = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(delayAvg))
```

Determine the number of distinct planes and flights by destination location.
```{r}
flights %>%
  group_by(dest) %>%
  summarize(planes = n_distinct(tailnum),
            flights = n())
```

Repeat but arrange from most to fewest planes.

## Tornado data

Let's review by using different data. The dataset was used to publish this recent [paper](http://myweb.fsu.edu/jelsner/PDF/Research/ElsnerEtAl2016.pdf). The data set through 2011 is available as a spreadsheet on my website.
```{r}
L = "http://myweb.fsu.edu/jelsner/temp/data/Tornadoes.txt"
Torn.df = read.table(L, header = TRUE)
dim(Torn.df)
head(Torn.df)
```

Each observation is a separate tornado. Variables include date, time, state, EF damage rating scale, fatalities, injuries, loss, start location, end location, length and width of path, and a few other things.

The start and end locations are given in decimal degrees of latitude and longitude. Missing locations are coded as 0.00. Path length is given in miles and path width is given in yards.

Reports are arranged chronologically starting with January 1950. The first tornado in the record is from January 3rd, 1950. It occurred in Missouri. 

There are quite a few tornadoes (56,221) so first make a tibble version of the data frame.
```{r}
Torn.df = tbl_df(Torn.df)
Torn.df
```

### Filter

Keep only tornadoes that occur in January.
```{r}
Torn.df %>%
  filter(MONTH == 1) %>%
  dim()
```

You read it as "Take the data frame `Torn.df` and filter it to remove all tornadoes except those occurring during the month of January."

You now have a filtered version of the original data frame consisting only of tornadoes occurring during the month of January. The function `dim()` on the filtered data frame or the function `length()` on one of the columns let's you know how many January tornadoes.

Multiple filtering conditions are joined together with a comma (or & for 'and').
```{r}
Torn.df %>%
  filter(MONTH == 1 & FSCALE >= 2)
```

The ampersand means 'and'. Filter by removing tornadoes except those occurring during January and with an damage rating of at least two.

You use the assignment statement to save the result of the filtering. The resulting object is a data frame and can be used by the function `ggplot()`.
```{r}
df = Torn.df %>%
  filter(MONTH == 6 & DAY == 16)
```

Here the functions `ggplot()` and `geom_point()` plot the tornado latitude on the vertical (y) axis against year on the horizontal (x) axis for all tornadoes that have occurred on June 16th.
```{r}
library(ggplot2)
ggplot(df, aes(x = YEAR, y = SLAT)) + 
  geom_point() + 
  geom_smooth(method = lm)
```

The function `aes()` with `x = YEAR` and `y = SLAT` puts the year on the horizontal axis and latitude on the vertical axis. The function `geom_point()` places points on the graph. The function `geom_smooth()` with argument `method = lm` adds a straightline fit through the points with a 95% confidence band about the line.

The graph indicates that tornado activity is shifting northward over time. There is large variability from one year to the next and it is just one day of the year so no definitive conclusions should be drawn.

### Arrange and select

List the tornadoes in descending order of fatalities and injuries.
```{r}
Torn.df %>%
  arrange(desc(FATALITIES), desc(INJURIES)) %>%
  select(DATE, FATALITIES, INJURIES)
```

The first tie on fatalities occurs with 58. The tie is broken by the number of injuries.

### Mutate

Add a column that is the path area in square meters. Select only the date, EF rating, and columns corresponding to the path length, width and area.
```{r}
df = Torn.df %>%
  mutate(Length = LENGTH * 1609,
         Width = WIDTH * .9144,
         Area = Length * Width * pi/4) %>%
  select(DATE, FSCALE, Length, Width, Area)
df
```

### Summarize

Compute the median tornado path length and width.
```{r}
df %>%
  summarize(mL = median(Length),
            mW = median(Width))
```

The median length is 805 meters and the median width is 37 meters.

### Commonalities

The functions `filter()`, `arrange()`, `select()`, `mutate()`, and `summarize()` have similar syntax. The first argument is always a data frame. Subsequent arguments describe what to do with it. You refer to columns in the data frame directly (without using `$`). The result is a new data frame. 

These properties make it easy to chain together multiple simple steps to achieve a complex result. These five functions provide the basis of a language of data manipulation.

At the most basic level, you can only alter a data frame in five useful ways: you can reorder the rows (arrange()), pick observations and variables of interest (filter() and select()), add new variables that are functions of existing variables (mutate()) or collapse many values to a summary (summarise()).

### Grouped operations

The verbs are useful, but they become powerful when you combine them with the idea of 'group by', repeating the operation individually on groups of observations within the dataset. 

You use the `group_by()` function to create a grouped object, which is a data frame divided by groups of rows. The verbs automatically work 'by group' when the input is a grouped object

In the following example, you group the data frame into `FSCALE` ratings and then count the number of tornadoes (`nT = n()`) and average the length (`mL = mean(Length)`) and width (`mW = mean(Width)`) for each group. Results are saved in a new data frame called `Paths`.
```{r}
Paths = Torn.df %>%
  mutate(Length = LENGTH * 1609,
         Width = WIDTH * .9144) %>%
  group_by(FSCALE) %>%
  summarise(nT = n(),
            mL = mean(Length),
            mW = mean(Width))
Paths
```

Remove observations without an F scale rating of -99.
```{r}
Paths = Paths %>%
  filter(FSCALE >= 0)
ggplot(Paths, aes(x = factor(FSCALE), y = mL)) +
  geom_bar(stat = "identity")
```

You use `summarise()` with aggregate functions, which take a vector of values, and return a single number. There are many useful functions in base R like `min()`, `max()`, `mean()`, `sum()`, `sd()`, `median()`, and `IQR()`. Other include:

* `n()`: number of observations in the current group.
* `n_distinct(x)`: number of unique values in x.

For example, we could use these to find the number of tornadoes by state and the number of months in which there was at least one tornado.
```{r}
Torn.df %>%
  group_by(STATE) %>%
  summarize(months = n_distinct(MONTH),
            nT = n())
```

You can also group by multiple variables. For example how many tornadoes have occurred by day of year?
```{r}
Torn.df %>%
  group_by(MONTH, DAY) %>%
  summarize(nT = n())
```

This is what statisticians call exploratory data analysis, or EDA for short. EDA is an interative cycle. You:

1. Generate questions about your data.
2. Search for answers by transforming, visualizing, and modeling your data.
3. Use what you learn to refine your questions and/or ask new ones.

Your goal during EDA is to develop and understanding of your data. Use questions as tools to guide your investigation. When you ask a question, the question focuses your attention on a specific part of your dataset and helps you decide which transformations to make. For example:

On average what state has the longest tornadoes? Hint: `group_by()`, `summarize()`, `arrange()`.

Please check out http://r4ds.had.co.nz/index.html for additional help.

## Problem Set # 3

Due Date: October 3, 2017
Total Points: 45

**1** The dataset *hflights* from the **hflights** package contains all 227,496 flights that departed Houston in 2011. Using the functions in the **dplyr** package

a. Create a new data frame from *hflights* containing only those flights that departed on September 11th of that year. (5)

b. How many flights departed on that day? (5)

c. Create a data frame with the first column being the tail number and the second being the number of departures from Houston the plane made that year sorted by most to least number of flights. (5)

**2** Answer the following questions using the tornado data set (http://myweb.fsu.edu/jelsner/temp/data/Tornadoes.txt).

a. Use the functions `group_by()` and `summarize()` to determine the number of tornadoes by each state. (5)

b. Create a new column with path length in meters. Create a new data frame by removing rows with EF damage rating below 3. Group by year and summarize the average path length. Make a graph with year on the x axis and average path length on the y axis. (10)

c. Compute the number of tornadoes with EF damage rating 1 or higher by year and graph the results. (10)

d. Create a data frame with the year in the first column and the total number of tornadoes in Kansas by year in the second column. List the first six rows of this new data frame. (5)
