---
title: "Data Munging Again"
author: "James B. Elsner"
date: September 13, 2018
output: 
  html_document:
    keep_md: true
---

"**You're doing it right if you get frustrated: if you're not frustrated, you're (probably) not stretching yourself mentally.**"---Hadley Wickham

Problem Set #2 is on Canvas. It's due next Thursday (September 20).

After today you will know additional ways to manipulate data frames. 

## Grammar for data

The **dplyr** package has functions ('verbs') that perform common operations on data frames. Selecting specific columns, filtering on rows, re-ordering rows, adding new columns, and summarizing data. Let's review using the `airquality` data frame.
```{r}
library(dplyr)
head(airquality)
dim(airquality)
```

### Select

The function `select()` selects variables by name. For example, create a new data frame containing only the month, day, and temperature columns from the `airquality` data frame.
```{r}
df <- airquality %>%
      select(Month, Day, Temp)
head(df)
```

### Filter

The function `filter()` filters observations based on specific values. Create a new data frame by filtering the rows keeping only the rows where temperature is at least 80 F and winds are less than 5 mph.
```{r}
df <- airquality %>% 
  filter(Temp >= 80 & Wind < 5)
head(df)
```

### Arrange

The function `arrange()` arranges (orders) rows by a particular column.
```{r}
airquality %>%
  arrange(Solar.R) %>%
  head()
```

String the verbs together.
```{r}
airquality %>%
  select(Solar.R, Wind, Temp) %>%
  filter(Temp > 90) %>%
  arrange(Temp, Solar.R)
```

### Mutate

The `mutate()` function adds new columns to the data frame. For example, create a new column called `TempC` which is the temperature in Celcius. Also create a column called `WindMS` which is the wind speed in meters per second.
```{r}
airquality %>%
  mutate(TempC = (Temp - 32) * 5/9,
         WindMS = Wind * .44704) %>%
  head()
```

On days when the temperature is below 60 F compute the apparent temperature based on wind chill and arrange from coldest to warmest apparent temperature.
```{r}
airquality %>%
  filter(Temp < 60) %>%
  mutate(TempAp = 35.74 + .6215 * Temp - 35.75 * Wind^.16 + .4275 * Temp * Wind^.16) %>%
  arrange(TempAp)
```

### Summarize

The `summarize()` (or `summarise()`) function allows reduces the data frame using summary statistics. To compute the average wind speed during July type
```{r}
airquality %>%
  filter(Month == 7) %>%
  summarize(Wavg = mean(Wind))
```

We've looked at several summary functions already including `sum()`, `sd()`, `min()`, `max()`, `var()`, `range()`, `median()`. Others include

Summary function  | Description
-----------------:|:-----------
`n()`             | Length of the column
`first()`         | First value of the column
`last()`          | Last value of the column
`n_distinct()`    | Number of distinct values

Find the maximum and median wind speed and maximum ozone concentration values during the month of May. Also determine the number of days during Mays.
```{r}
airquality %>%
  filter(Month == 5) %>%
  summarize(Wmax = max(Wind),
            Wmed = median(Wind),
            OzoneMax = max(Ozone, na.rm = TRUE),
            NumDays = n())
```

### Grouping

What if we want to summarize separately for each month? 

We use the `group_by()` function. We split the data frame by some variable (e.g., `Month`), apply a function to the individual data frames, and then combine the output.

Find the highest ozone concentration for each month.
```{r}
airquality %>%
  group_by(Month) %>%
  summarize(OzoneMax =  max(Ozone, na.rm = TRUE),
            NumDays = n())
```

Find the average ozone concentration when temperatures are above and below 70 F.
```{r}
airquality %>%
  group_by(Temp >= 70) %>%
  summarize(OzoneAvg =  mean(Ozone, na.rm = TRUE),
            NumDays = n())
```

On average ozone concentrations are higher on warm days (Temp >= 70 F) days. That is to say mean ozone concentration depends on temperature, where the word 'depends' implies statistical dependency.

The mean is a model for the data. The statistical dependency of the mean implies that a model for ozone concentration will likely be improved by including temperature as an explanatory variable.

The important **dplyr** verbs are

Verb          | Description
-------------:|:-----------
`select()`    | selects columns; pick variables by their names
`filter()`    | filters rows; pick observations by their values
`arrange()`   | re-orders the rows
`mutate()`    | creates new columns; create new variables with functions of existing variables
`summarize()` | summarizes values; collapse many values down to a single summary
`group_by()`  | allows operations to be grouped

The syntax of the verb functions in the **dplyr** package are all the same:

* The first argument is a data frame. This argument is implicit when using the `%>%` operator.
* The subsequent arguments describe what to do with the data frame. We refer to columns in the data frame directly (without using `$`).
* The result is a new data frame

These properties make it easy to chain together many simple lines of code to do something complex.

The five functions provide the basis of a grammar for data. At the most basic level, we can only alter a data frame in five useful ways: we can reorder the rows (`arrange()`), pick observations and variables of interest (`filter()` and `select()`), add new variables that are functions of existing variables (`mutate()`), or collapse many values to a summary (`summarise()`).

## New York City flight data

Let's apply these functions to a larger set of data. The data frame is in the package **nycflights13**. The data set is called `flights`. The following notes are from https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html
```{r}
library(nycflights13)
dim(flights)
```

The data contains all 336,776 flights that departed NYC in 2013 and comes from the U.S. Bureau of Transporation Statistics. More information is available by typing `?nycflights13`.

The object `flights` is a tibble (tabled data frame). When we have a large data frame it is useful to make it a tibble so that we don't accidently print it to the screen.

For example, `airquality` is not a tibble. So when we type the name of the data frame the entire data frame scrolls in the console. To change a data frame to a tibble, use the function `tbl_df()`.
```{r}
airquality
airquality <- tbl_df(airquality)
airquality
```

Returning to the NYC flights data.
```{r}
flights
```

The function `filter()` selects a set of rows in a data frame 

How would we select all flights occurring on February 1st?
```{r}
flights %>%
  filter(month == 2 & day == 1)
```

The function `arrange()` works like `filter()` except that it reorders the rows. If we provide more than one column name, each additional column is used to break ties in the values of the preceding columns.

How would we arrange all flights in descending order of departure delay?
```{r}
flights %>%
  arrange(desc(dep_delay))
```

Often we work with large datasets with many columns but only a few are of interest. The function `select()` allows us to zoom in on an interesting subset of the columns.

How would we create a data frame containing only the dates, carrier, and flight numbers?
```{r}
df <- flights %>%
  select(year:day, carrier, flight)
df
```

We add new columns with the function `mutate()`.

Compute the time gained in flight by subtracting the departure delay (minutes) from the arrival delay.
```{r}
flights %>%
  mutate(gain = arr_delay - dep_delay) %>%
  select(year:day, carrier, flight, gain) %>%
  arrange(desc(gain))
```

Determine the mean departure delay.
```{r}
flights %>%
  summarize(avgDelay = mean(dep_delay, na.rm = TRUE))
```

Note that if there are missing values in a vector the function `mean()` needs to include the argument `na.rm = TRUE` otherwise the output will be `NA`.

```{r}
y <- c(5, 6, 7, NA)
mean(y)
mean(y, na.rm = TRUE)
```

### Randomly sample rows

We use `sample_n()` and `sample_frac()` to take random sample of rows from the data frame. Take a random sample of five rows from the flights data frame.
```{r}
flights %>%
  sample_n(5)
```

Take a random sample of 1% of the rows.
```{r}
flights %>%
  sample_frac(.01)
```

Use the argument `replace = TRUE` to perform a bootstrap sample. More on this later.

The verbs are powerful when we apply them to groups of observations within a data frame. This is done with the function `group_by()`. Determine the average arrival delay by airplane (tail number).
```{r}
flights %>%
  group_by(tailnum) %>%
  summarize(delayAvg = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(delayAvg))
```

Determine the number of distinct planes and flights by destination location.
```{r}
flights %>%
  group_by(dest) %>%
  summarize(planes = n_distinct(tailnum),
            flights = n())
```

Repeat but arrange from most to fewest planes.

Data munging is part of data science. It is an interative cycle:

1. Generate questions about our data.
2. Search for answers by transforming, visualizing, and modeling the data.
3. Use what we learn to refine our questions and/or ask new ones.

The goal is to develop understanding. We use questions as tools to guide our investigation. When we ask a question, the question focuses our attention on a specific part of our dataset and helps us decide what to do. For example: On average what state has the longest tornadoes? Hint: `group_by()`, `summarize()`, `arrange()`.

For additional practice please check out http://r4ds.had.co.nz/index.html.

Cheat sheet https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

## Problem Set #2