---
title: "Data Munging Again"
author: "James B. Elsner"
date: September 13, 2018
output: 
  html_document:
    keep_md: true
---

"**You're doing it right if you get frustrated: if you're not frustrated, you're (probably) not stretching yourself mentally.**"---Hadley Wickham

Problem Set #2 is on Canvas. It's due next Thursday (September 20).

After today you will be better at manipulating data frames. 

## Grammar

The syntax of the verb functions in the **dplyr** package are all the same:

* The first argument is a data frame. This argument is implicit when using the `%>%` operator.
* The subsequent arguments describe what to do with the data frame. We refer to columns in the data frame directly (without using `$`).
* The result is a new data frame

These properties make it easy to chain together many simple lines of code to do something complex.

The five functions provide the basis of a grammar for data. At the most basic level, we can only alter a data frame in five useful ways: we can reorder the rows (`arrange()`), pick observations and variables of interest (`filter()` and `select()`), add new variables that are functions of existing variables (`mutate()`), or collapse many values to a summary (`summarise()`).

## Tornado data

Let's continue using data from my research project. The dataset was used to publish this recent [paper](http://myweb.fsu.edu/jelsner/PDF/Research/ElsnerEtAl2016.pdf). The data set through 2011 is available as a spreadsheet on my website.
```{r}
L = "http://myweb.fsu.edu/jelsner/temp/data/Tornadoes.txt"
Torn.df = read.table(L, header = TRUE)
dim(Torn.df)
head(Torn.df)
```

Each observation is a separate tornado. Variables include date, time, state, EF damage rating scale, fatalities, injuries, loss, start location, end location, length and width of path, and a few other things.

The start and end locations are given in decimal degrees of latitude and longitude. Missing locations are coded as 0.00. Path length is given in miles and path width is given in yards.

Reports are arranged chronologically starting with January 1950. The first tornado in the record is from January 3rd, 1950. It occurred in Missouri. 

There are quite a few tornadoes (56,221) so first make a tibble version of the data frame.
```{r}
Torn.df <- tbl_df(Torn.df)
Torn.df
```

### Filter

Keep only tornadoes that occur during the month of January.
```{r}
Torn.df %>%
  filter(MONTH == 1) %>%
  dim()
```

We read it as "Take the data frame `Torn.df` and filter it to remove all tornadoes except those occurring during the month of January."

You now have a filtered version of the original data frame consisting only of tornadoes occurring during the month of January. The function `dim()` on the filtered data frame or the function `length()` on one of the columns let's you know how many January tornadoes.

Multiple filtering conditions are joined together with a comma (or & for 'and').
```{r}
Torn.df %>%
  filter(MONTH == 1 & FSCALE >= 2)
```

The ampersand means 'and'. Filter by removing tornadoes except those occurring during January and with an damage rating of at least two.

You use the assignment statement to save the result of the filtering. The resulting object is a data frame and can be used by the function `ggplot()`.
```{r}
df <- Torn.df %>%
  filter(MONTH == 6 & DAY == 16)
```

Here the functions `ggplot()` and `geom_point()` plot the tornado latitude on the vertical (y) axis against year on the horizontal (x) axis for all tornadoes that have occurred on June 16th.
```{r}
library(ggplot2)
ggplot(df, aes(x = YEAR, y = SLAT)) + 
  geom_point() + 
  geom_smooth(method = lm)
```

The function `aes()` with `x = YEAR` and `y = SLAT` puts the year on the horizontal axis and latitude on the vertical axis. The function `geom_point()` places points on the graph. The function `geom_smooth()` with argument `method = lm` adds a straightline fit through the points with a 95% confidence band about the line.

The graph indicates that tornado activity is shifting northward over time. There is large variability from one year to the next and it is just one day of the year so no definitive conclusions should be drawn.

### Arrange and select

List the tornadoes in descending order of fatalities and injuries.
```{r}
Torn.df %>%
  arrange(desc(FATALITIES), desc(INJURIES)) %>%
  select(DATE, FATALITIES, INJURIES)
```

The first tie on fatalities occurs with 58. The tie is broken by the number of injuries.

### Mutate

Add a column that is the path area in square meters. Select only the date, EF rating, and columns corresponding to the path length, width and area.
```{r}
df <- Torn.df %>%
  mutate(Length = LENGTH * 1609,
         Width = WIDTH * .9144,
         Area = Length * Width * pi/4) %>%
  select(DATE, FSCALE, Length, Width, Area)
df
```

### Summarize

Compute the median tornado path length and width.
```{r}
df %>%
  summarize(mL = median(Length),
            mW = median(Width))
```

The median length is 805 meters and the median width is 37 meters.

### Commonalities

The functions `filter()`, `arrange()`, `select()`, `mutate()`, and `summarize()` have similar syntax. The first argument is always a data frame. Subsequent arguments describe what to do with it. You refer to columns in the data frame directly (without using `$`). The result is a new data frame. 

These properties make it easy to chain together multiple simple steps to achieve a complex result. These five functions provide the basis of a language of data manipulation.

At the most basic level, you can only alter a data frame in five useful ways: you can reorder the rows (arrange()), pick observations and variables of interest (filter() and select()), add new variables that are functions of existing variables (mutate()) or collapse many values to a summary (summarise()).

### Grouped operations

The verbs are useful, but they become powerful when you combine them with the idea of 'group by', repeating the operation individually on groups of observations within the dataset. 

You use the `group_by()` function to create a grouped object, which is a data frame divided by groups of rows. The verbs automatically work 'by group' when the input is a grouped object

In the following example, you group the data frame into `FSCALE` ratings and then count the number of tornadoes (`nT = n()`) and average the length (`mL = mean(Length)`) and width (`mW = mean(Width)`) for each group. Results are saved in a new data frame called `Paths`.
```{r}
Paths <- Torn.df %>%
  mutate(Length = LENGTH * 1609,
         Width = WIDTH * .9144) %>%
  group_by(FSCALE) %>%
  summarise(nT = n(),
            mL = mean(Length),
            mW = mean(Width))
Paths
```

Remove observations without an F scale rating of -99.
```{r}
Paths = Paths %>%
  filter(FSCALE >= 0)
ggplot(Paths, aes(x = factor(FSCALE), y = mL)) +
  geom_bar(stat = "identity")
```

We use `summarise()` with aggregate functions, which take a vector of values, and return a single number. There are many useful functions in base R like `min()`, `max()`, `mean()`, `sum()`, `sd()`, `median()`, and `IQR()`. Other include:

* `n()`: number of observations in the current group.
* `n_distinct(x)`: number of unique values in x.

For example, we could use these to find the number of tornadoes by state and the number of months in which there was at least one tornado.
```{r}
Torn.df %>%
  group_by(STATE) %>%
  summarize(months = n_distinct(MONTH),
            nT = n())
```

You can also group by multiple variables. For example how many tornadoes have occurred by day of year?
```{r}
Torn.df %>%
  group_by(MONTH, DAY) %>%
  summarize(nT = n())
```

This is what statisticians call exploratory data analysis, or EDA for short. EDA is an interative cycle. You:

1. Generate questions about your data.
2. Search for answers by transforming, visualizing, and modeling your data.
3. Use what you learn to refine your questions and/or ask new ones.

Your goal during EDA is to develop and understanding of your data. Use questions as tools to guide your investigation. When you ask a question, the question focuses your attention on a specific part of your dataset and helps you decide which transformations to make. For example:

On average what state has the longest tornadoes? Hint: `group_by()`, `summarize()`, `arrange()`.

Please check out http://r4ds.had.co.nz/index.html for additional help.


## New York City flight data

Let's apply these functions to an even larger set of data. The data frame is in the package **nycflights13**. The data set is called `flights`. The notes follow closely from https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html
```{r}
library(nycflights13)
dim(flights)
```

The data contains all 336,776 flights that departed NYC in 2013 and comes from the U.S. Bureau of Transporation Statistics. More information is available by typing `?nycflights13`.

The object `flights` is a tibble (tabled data frame). When you are dealing with a large data frame it is useful to make it a tibble so that you don't accidently print it to the screen.

For example, `airquality` is not a tibble. So when you type the name of the data frame the entire data frame scrolls in the console. To change a data frame to a tibble, use the function `tbl_df()`.
```{r}
airquality
airquality = tbl_df(airquality)
airquality
```

Returning to the NYC flights data.
```{r}
flights
```

The function `filter()` allows you to select a subset of rows in a data frame 

How would you extract all flights occurring on February 1st?
```{r}
flights %>%
  filter(month == 2 & day == 1)
```

The function `arrange()` works like `filter()` except that it reorders the rows. If you provide more than one column name, each additional column will be used to break ties in the values of the preceding columns.

How would you arrange all flights in descending order of departure delay?
```{r}
flights %>%
  arrange(desc(dep_delay))
```

Often you work with large datasets with many columns but only a few may be of interest. The function `select()` allows you to zoom in on an interesting subset of the columns.

How would you create a data frame containing only the dates, carrier, and flight numbers?
```{r}
df <- flights %>%
  select(year:day, carrier, flight)
df
```

You add new columns with the function `mutate()`.

Compute the time gained in flight by subtracting the departure delay (minutes) from the arrival delay.
```{r}
flights %>%
  mutate(gain = arr_delay - dep_delay) %>%
  select(year:day, carrier, flight, gain) %>%
  arrange(desc(gain))
```

Determine the mean departure delay.
```{r}
flights %>%
  summarize(avgDelay = mean(dep_delay, na.rm = TRUE))
```

### Randomly sample rows

You use `sample_n()` and `sample_frac()` to take random sample of rows from your data frame. Take a random sample of five rows from the flights data frame.
```{r}
flights %>%
  sample_n(5)
```

Take a random sample of 1% of the rows.
```{r}
flights %>%
  sample_frac(.01)
```

Use the argument `replace = TRUE` to perform a bootstrap sample. More on this later.

The verbs are powerful when you apply them to groups of observations within a data frame. This is done with the function `group_by()`. Determine the average arrival delay by airplane (tail number).
```{r}
flights %>%
  group_by(tailnum) %>%
  summarize(delayAvg = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(delayAvg))
```

Determine the number of distinct planes and flights by destination location.
```{r}
flights %>%
  group_by(dest) %>%
  summarize(planes = n_distinct(tailnum),
            flights = n())
```

Repeat but arrange from most to fewest planes.


Use google to search for data sets.
https://toolbox.google.com/datasetsearch


EDA is an interative cycle. You:

1. Generate questions about your data.
2. Search for answers by transforming, visualizing, and modeling your data.
3. Use what you learn to refine your questions and/or ask new ones.

Your goal is to develop an understanding of your data. Use questions as tools to guide your investigation. When you ask a question, the question focuses your attention on a specific part of your dataset and helps you decide which transformations to make. For example: On average what state has the longest tornadoes? Hint: `group_by()`, `summarize()`, `arrange()`.

```{r}
Torn.df %>%
  group_by(STATE) %>%
  summarize(mL = mean(LENGTH)) %>%
  arrange(desc(mL))
```

For additional practice please check out http://r4ds.had.co.nz/index.html.

## Problem Set #2
