---
title: "Making Graphs"
author: "James B. Elsner"
date: September 18, 2018
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

"**Good coding style is like using correct punctuation. You can manage without it, but it sure makes things easier to read.**"---Hadley Wickham

After today's class you will know how to make some common graphs in R using the functions from the **ggplot2** package.

## Grammer for graphs

The **ggplot2** package has functions for making nice graphs. The 'gg' stands for the 'Grammar of Graphics,' a theory of how to create a graphics system from Leland Wilkinson.

The grammar specifies how a plot translates data to attributes and geometric objects. Attributes are things like color, shape, and size. Geometric objects are things like points, lines, bars, and polygons. A plot is drawn using a coordinate system. The type of plot depends on the geometric object, which is specified as a function. 

The function names for geometric objects in **ggplot2** begin with `geom_`. For example, to create a scatter plot of points the `geom_point()` function is used. The plot may contain manipulations of the data (like `geom_histogram()`). Similar plots on subsets of the data are mading using the `facet_` functions (similar to the `group_by()` function).

Load the **ggplot2** package.
```{r}
library(ggplot2)
```

NOTE: Always make sure you have a call to the package load function (`library()`) in your code before you knit.

### Bar chart

One of the simplest graphs is the bar chart showing the number of cases on each group. Although the graph is simple, it involves a manipulation of the data.

Let's start with a bar plot of the hurricane counts. Read the hurricane data from the file on my website and list the first six rows.
```{r}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/US.txt"
H <- read.table(loc, 
                header = TRUE)
head(H)
```

Recall that the function `table()` creates a frequency table from the landfall counts. For example, we create a frequency of years with n number of hurricanes.
```{r}
table(H$All)
```

The number of cases for each count is tallied and displayed below the count. There were 36 cases of 0 hurricanes. Each case is a year.

The function `geom_bar()` creates a bar chart of this frequency table.
```{r}
ggplot(data = H) + 
  geom_bar(mapping = aes(x = All))
```

With **ggplot2**, we begin a plot with the function `ggplot()`. `ggplot()` creates a coordinate system that we can add layers to. The first argument of `ggplot()` is the dataset to use in the graph. So `ggplot(data = H)` creates an empty graph.

We complete our graph by adding one or more layers to `ggplot()`. The function `geom_bar()` adds a layer of bars to our plot, which creates a bar chart. **ggplot2** comes with many `geom_` functions that each add a different type of layer to a plot.

Each `geom_X` function takes a mapping argument. This defines how variables in our dataset are mapped to visual properties. The mapping argument is always paired with `aes()` function, and the `x` argument of `aes()` specifies which variables to map to the x axes, in this case `All`. `ggplot()` looks for the mapped variable in the data argument, in this case, `H`.

The function `geom_bar()` tables the counts and then maps the number of cases to bars with the bar height proportional to the number of cases. Here the number of cases is the number of years with that many hurricanes.

The functions are applied in order (`ggplot()` comes before `geom_bar()`) and are linked with the addition `+` symbol. In this way we can think of the functions as layers in a GIS.

The bar chart contains the same information as displayed by the function `table()`. The y-axis label is 'count' and x-axis label is the column name. 

Repeat this time using Florida hurricane counts. The annual number of Florida hurricanes by year is given in column `FL` in the data frame `H`.
```{r}
H$FL
```

The `geom_bar()` function tables these numbers and plots the frequency as a bar.
```{r}
ggplot(data = H) + 
  geom_bar(mapping = aes(x = FL)) +
  xlab("Number of Florida Hurricanes (1851-2016)") +
  ylab("Number of Years")
```

Here axes labels are placed on the plot with the functions `ylab()` and `xlab()`. With this type of 'layering' it's easy to go from data on the web to a publishable plot.

### Beer preference

Thirty graduate students are surveyed about their beer drinking preferences. Categories are (1) domestic can, (2) domestic bottle, (3) microbrew, and (4) import. To make a bar chart first create the data as a character vector and then change the vector to a data frame.
```{r}
beer <- c(rep('domestic bottle', 4), 
          rep('domestic can', 12), 
          rep('import', 5), 
          rep('microbrew', 9))
beerPref.df <- as.data.frame(beer)
```

Use the function `str()` to see the column type in the data frame.
```{r}
str(beerPref.df)
```

There is a single column in the data frame with the name `beer`. It is a factor variable with 4 levels one for each type of beer choice. A factor is a categorical vector. It looks like a character but it can be ordered. This is important when factors are used in statistical models.

Create a table.
```{r}
table(beerPref.df$beer)
```

Create a bar chart and specify the axis labels.
```{r}
ggplot(data = beerPref.df) + 
  geom_bar(mapping = aes(beer)) +
  xlab("Beer Preference") + 
  ylab("Number of Students")
```

This is a good start but a few improvements should be made.

1. The bar order is alphabetical from left to right. This is the default ordering for factor variables created from character vectors. It is much easier to make comparisons if frequency determines the order.

To change the order on the bar chart specify the factor levels on the vector beer.
```{r}
beer <- factor(beer, 
               levels = c("domestic can", "microbrew", "import", "domestic bottle"))
beerPref.df <- data.frame(beer)
```

Now remake the bar chart.
```{r}
ggplot(beerPref.df) + 
  geom_bar(mapping = aes(beer)) +
  xlab("Beer Preference") + 
  ylab("Number of Students")
```

2. The vertical axis tic labels are fractions. Since the bar heights are counts (integers) the tic labels also should be integers.

To override this default you add a new y-axis layer. The layer is the function `ylim()` where you indicate the lower and upper limits of the axis with the concatenate (`c()`) function. Now remake the bar chart.
```{r}
ggplot(beerPref.df) + 
  geom_bar(mapping = aes(beer)) +
  xlab("Beer Preference") + 
  ylab("Number of Students") +
  ylim(c(0, 15))
```

Now the chart is publishable and this schema works for print media. Options exist for changing the look of the plot for digital media include, colors, orientation, background, etc.

For example to change the bar color use the `fill =` argument in the function `geom_bar()`. To change the orientation of the bars use the layer function `coord_flip`, and to change the background use the layer function `theme_minimal()`

We make changes to the look of the plot with additional layers.
```{r}
ggplot(beerPref.df) + 
  geom_bar(mapping = aes(beer), fill = "orange2") +
  xlab("Beer Preference") + 
  ylab("Number of Students") +
  ylim(c(0, 15)) +
  coord_flip() +
  theme_minimal()
```

Note that the `fill =` is used on the variable named in the `aes()` function but it is specified outside the `aes()` function.

Available colors include 
```{r eval=FALSE}
colors()
```

It's a bit more complicated to automatically change the bar order based on the frequency of occurrence. Here is an example for storm intensity of tropical cyclones during 2017 (last year). First create two vectors one numeric containing the minimum pressures (millibars) and the other character containing the storm names and combine them into a data frame.
```{r}
minP <- c(990, 1007, 992, 1007, 1005, 981, 967, 938, 914, 938, 972, 971)
name <- c("Arlene", "Bret", "Cindy", "Don", "Emily", "Franklin", "Gert", 
         "Harvey", "Irma", "Jose", "Katia", "Lee")
df <- data.frame(name, minP)
```

Next reorder the names based on minimum pressure. This is done with the `reorder()` function where the first argument is a character vector and the second argument is a vector of the same length but determines the order of the levels.
```{r}
library(dplyr)
df <- df %>%
         mutate(name = reorder(name, minP))
ggplot(df) + 
  geom_bar(mapping = aes(x = name, y = minP), stat = "identity") +
  ylab("Minimum Sea Level Pressure [mb]") 
```

### Histogram

The histogram is similar to the bar chart except the histogram uses bars to indicate frequency (or proportion) over an interval of continuous values. For instance, with continuous values the function `table()` is not useful.
```{r}
x <- rnorm(10)
table(x)
```

So neither is a bar plot.

A histogram is made as follows: First a contiguous collection of disjoint intervals, called bins, covering the range of data points is chosen.  "Disjoint" means no overlap, so the intervals look like (a,b] or [a,b). The interval (a,b] means the interval contains all the values from a to b including b but not a, whereas the interval [a,b) means the interval contains all the values from a to b including a but not b.

Second, the number of data values in each of these intervals is counted. Third, a bar is drawn above the interval so that the area of the bar is proportional to the frequency. If the intervals defining the bins have the same width, then the height of the bar is proportional to the frequency (the number of values inside the interval).

Let's return to the Florida precipitation data.
```{r}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/FLprecip.txt"
FLp <- read.table(loc, 
                  na.string = "-9.900", 
                  header = TRUE)
str(FLp)
```

Columns in the data frame `FLp` are months (variables) and rows are years. `Year` is an integer (int) vector and the months are numeric (num) vectors. Create a histogram of May precipitation.
```{r}
ggplot(data = FLp) + 
  geom_histogram(mapping = aes(May))  +
  xlab("May Precipitation in Florida (in)") 
```

By default the function `geom_histogram()` picks 30 bins. Since there are only 118 May values many of the bins have fewer than 5 values.

When we make a histogram we will almost certainly need to vary the number of bins before deciding on a final plot. This can be done with the `bins =` or `binwidth =` argument. For example, the look of the histogram is improved by halving the default number of bins.
```{r}
ggplot(data = FLp) + 
  geom_histogram(mapping = aes(May), bins = 15)  +
  xlab("May Precipitation in Florida (in)") 
```

It looks even better by decreasing the number of bins to 11.
```{r}
ggplot(FLp) + 
  geom_histogram(mapping = aes(May), bins = 11, fill = "green3")  +
  xlab("May Precipitation in Florida (in)") +
  ylab("Number of Years")
```

Here the fill argument is used to change color and a `ylab()` layer is added to make the y-axis label more concise.

The `geom_rug()` layer adds the location of the data values as tic marks just above the horizontal axis. And the `color = "gray"` is the color of the bin boundaries.
```{r}
ggplot(FLp) + 
  geom_histogram(mapping = aes(May), bins = 11, fill = "green3", color = "gray")  +
  xlab("May Precipitation in Florida (in)") +
  ylab("Number of Years") +
  geom_rug()
```

### Density plot

A density plot is a smoothed histogram with units of probability on the vertical axis. It's motivated by the fact that for a continuous variable, the probability that X takes on any particular value x is 0 so we need a range of values over which we can define a probability.

The probability density answers the question, what is the chance that X falls in a small interval. This chance varies depending on where X is located (e.g., near the middle of the distribution the chance is highest).
```{r}
ggplot(FLp) +
  geom_density(mapping = aes(May)) +
  xlab("May Precipitation in Florida (in)")
```

The probability axis (vertical) is the integral (average) chance that rainfall will take on a value along the horizontal axis within a given small interval. The size of the interval is determined by the bandwidth (`bw =`).
```{r}
ggplot(FLp) +
  geom_density(mapping = aes(May), bw = 1) +
  xlab("May Precipitation in Florida (in)")
```

The probability scale depends on the data units so it is often tricky to interpret. Instead `geom_freqpoly()` produces a density-like graph where the units on the y-axis are counts as with the histogram.
```{r}
ggplot(FLp) + 
  geom_freqpoly(mapping = aes(May), color = "green3", binwidth = 1) +
  xlab("May Precipitation in Florida (in)") +
  geom_rug()
```

### Box plot

The box plot graphs five summary statistics including the min, max, 1st & 3rd quartiles, and the median. The easiest way to create a box plot is to use the function `boxplot()`.
```{r}
boxplot(FLp$May)
```

The function `boxplot()` is from the base **graphics** package. It is not a **ggplot2** function. Others from this package include `hist()` for histograms and `plot()` for scatter plots. 

The base graphics lets you manipulate every detail of a graph, which is great for customization. For example:
```{r}
boxplot(FLp$May, 
        ylab = "May Precipitation in FL (in)")
f <- fivenum(FLp$May)
text(rep(1.3, 5), f, labels = c("Minimum", "1st Quartile", 
                                "Median", "3rd Quartile",
                                "Maximum"))
text(1.3, 7.792, labels = "Last Value Within\n 1.5xIQR Above 3rd Q")
```

The box plot illustrates the five numbers graphically. The median is the line through the box. The bottom and top of the box are the 1st and 3rd quartile values. Whiskers extend vertically from the box downward toward the minimum and upward toward the maximum. 

If values extend beyond 1.5 times the interquartile range (either above or below the corresponding quartile) the whisker is truncated at the last value within the range and points are used to indicate outliers.

To make a box plot using the function `ggplot()` we need a dummy variable for the `x` argument in the function `aes()`. This is done with `x = ""`.
```{r}
ggplot(FLp) + 
  geom_boxplot(mapping = aes(x = "", y = May)) +
  xlab("") + 
  ylab("May Precipitation in Florida (in)")
```

### Side-by-side box plots

Suppose we want to show box plots for each month. In this case we make the x argument in the `aes()` the name of a column containing the vector of month names. We need to first turn the data frame from its native 'wide' format to a **gplot2** friendly 'long' format.

#### Wide versus long data formats

Wide format data is called 'wide' because it typically has a lot of columns which stretch across the page or our computer screen. Long format data is called 'long' because it typically has fewer columns but preserves all the same information. In order to do that, it has to be longer.

We are most familiar with wide format data. It is convenient for data entry. It lets us see more of the data at one time.

The long format is less familiar. It corresponds to the relational model for storing data used by most modern databases like SQL.

Use `gather()` from the **tidyr** package to turn the wide data frame into a long data frame.
```{r}
library(tidyr)
head(FLp)
FLpL <- FLp %>%
  gather(key = Year,
         value = "Precipitation",
         factor_key = TRUE) %>%
  rename(Month = Year)
head(FLpL)
```

The `gather()` function takes all the values as measured except those named in `key = ` argument. All variables are measured (precipitation in units of inches) except `Year`. `Year` is a vector identifying the month. 

The long data frame lists the key variable names as the first column taking the column names as character strings starting with `Jan`. To preserve the order of the columns (e.g., January comes before February, etc) we specify the `factor_key = TRUE`. We use `rename()` to give the key variable the correct name. 

Then to create the box plot specify that the x-axis be the key variable (here `Month`) and the y-axis to be the measured variable (here `Precipitation`).
```{r}
ggplot(FLpL) + 
  geom_boxplot(mapping = aes(x = Month, y = Precipitation)) +
  ylab("Precipitation (in)")
```

This is a climograph.

Each `geom_` function is a layer. Data for the layer is specified in the function `ggplot()` with the data frame argument and the `aes()` function. To add another layer to the plot with different data you specify the data within the geom function. For example, lets repeat the climograph of monthly precipitation highlighting the month of May.

We add another `geom_boxplot()` layer and specify a subset of the data using the subset `[]` operator.
```{r}
ggplot(FLpL, aes(x = Month, y = Precipitation)) + 
  geom_boxplot() +
  ylab("Precipitation (in)") +
  geom_boxplot(data = FLpL[FLpL$Month == "May", ], 
               aes(x = Month, y = Precipitation), 
               fill = "green")
```

Practice: Since this is a common graph for climatologists lets practice with another set of data. The data are monthly values of the Southern Oscillation Index (indicator of El Ni\~no) and are available on my website with the name `SOI.txt`. Make a climograph from the monthly SOI data. Here I put the functions in the proper order but I left out some details that I want you to fill in.
```{r eval=FALSE}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/"
SOI <- read.table(file = ,
                  header = TRUE)
SOIL <- SOI %>%
  gather(key = , value = Index) %>%
  rename()
ggplot(SOIL) + 
  geom_boxplot(mapping = aes(x = , y = )) +
  ylab("SOI (s.d.)")
```

### Scatter plot

The workhorse of statistical graphs is the scatter plot. It's a graph of the values of one variable against the values of the other as points $(x_i, y_i)$ in a Cartesian plane. It's used to show the relationship between two numeric variables. 

For example, to show the relationship between April and September values of Florida precipitation we type
```{r}
ggplot(FLp) + 
  geom_point(mapping = aes(x = Apr, y = Sep)) + 
  xlab("April Precip. (in)") + 
  ylab("September Precip. (in)")
```

The plot shows that dry Aprils tend to be followed by dry Septembers and wet Aprils tend to be followed by wet Septembers. We observe there is a direct (or positive) relationship between the two variables although the points are scattered widely indicating the relationship is weak.

If our goal is to model the relationship, we plot the dependent variable (the variable we are interested in modeling) on the vertical axis. Here we put the September values on the vertical axis since a predictive model would use April values to forecast September values.

If the points have a natural ordering then we use the `geom_line()` function. For example, to plot the September precipitation values as a time series type
```{r}
ggplot(FLp) + 
  geom_line(mapping = aes(x = Year, y = Sep)) + 
  xlab("Year") + 
  ylab("September Precip. (in)")
```

Precipitation values fluctuate from one September to the next, but there does not appear to be a long-term trend. With time series data it is better to connect the values with lines rather than use points unless values are missing.

Practice: Create a plot of the May values of the North Atlantic oscillation (NAO) with Year on the horizontal axis. Add appropriate axis labels.
```{r eval=FALSE}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/NAO.txt"
NAO <- read.table(file = loc, header = TRUE)
ggplot(NAO, aes(x = Year, y = May)) + 
  geom_line() + 
  xlab("Year") + 
  ylab("North Atlantic Oscillation (s.d.)")
```

### More than two variables

Let's return to the `mpg` data frame. The data frame contains different automobiles by who made it, the model, engine size, mileage, class, etc.
```{r}
names(mpg)
```

Let's start with a scatter plot showing highway mileage on the vertical axis and engine size on the horizontal axis.
```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

We can add a third variable, like `class`, to a two dimensional scatterplot by mapping it to an aesthetic. An aesthetic is a visual property of the objects in our plot. Aesthetics include things like the size, the shape, or the color of our points. We can display a point in different ways by changing the levels of its aesthetic properties (e.g., changing the level by size, color, type).

For example, you map the colors of our points to the `class` variable to reveal the class of each car.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

To map an aesthetic to a variable, associate the name of the aesthetic to the name of the variable inside `aes()`. Note in the previous plot `color =` was specified outside `aes()`. This is called 

`ggplot()` will automatically assign a unique level of the aesthetic (here a unique color) to each unique value of the variable, a process known as scaling. `ggplot()` will also add a legend that explains which levels correspond to which values.

The colors show that many of the unusual points are two-seater cars. Sports cars have large engines like SUVs and pickup trucks, but small bodies like midsize and compact cars, which improves their gas mileage.

For each aesthetic, we use `aes()` to associate the name of the aesthetic with a variable to display. The `aes()` function gathers each of the aesthetic mappings used by a layer and passes them to the laye's mapping argument. The syntax highlights a useful insight about `x` and `y`: the x and y locations of a point are themselves aesthetics, visual properties that we can map to variables to display information about the data.

Once we map an aesthetic, the **ggplot2** takes care of the rest. It selects a reasonable scale to use with the aesthetic, and it constructs a legend that explains the mapping between levels and values. For `x` and `y` aesthetics, **ggplot2** does not create a legend, instead it creates an axis line with tick marks and a label. The axis line acts as a legend; it explains the mapping between locations and values.

Question: What is wrong with this code? Why aren't the points blue?
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

### Facets

One way to add additional variables is with aesthetics. Another way, particularly useful for categorical variables, is to split our plot into facets, subplots that each display one subset of the data.

To facet a plot by a single variable, use `facet_wrap()`. The first argument of `facet_wrap()` should be a formula, which we create with `~` (tilde) followed by a variable name (here 'formula' is the name of a data structure in R, not a synonym for 'equation'). The variable that we pass to `facet_wrap()` should be discrete.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

To facet a plot on the combination of two variables, add `facet_grid()` to the plot call. The first argument of `facet_grid()` is also a formula. This time the formula should contain two variable names separated by a `~` with the first variable named varying in the vertical direction and the second varying in the horizontal direction. 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

Here `drv` refers to the drivetrain: front-wheel (`f`), rear-wheel (`r`) or 4-wheel (`4`).

### Geometric objects

A 'geom' is the geometrical object that a plot uses to represent data. People often describe plots by the type of geom that the plot uses. For example, bar charts use bar geoms, line charts use line geoms, boxplots use boxplot geoms, and so on. Scatterplots break the trend; they use the point geom. We can use different geoms to plot the same data.

To change the geom in our plot, change the geom function that we add to `ggplot()`. For instance,

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

And
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

Every geom function in **ggplot2** takes a mapping argument. However, not every aesthetic works with every geom. We could set the shape of a point, but we can't set the 'shape' of a line. On the other hand, we can set the line type of a line. `geom_smooth()` draws a different line, with a different linetype, for each unique value of the variable that we map to `linetype`.
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

Here `geom_smooth()` separates the cars into three lines based on their `drv` value, which describes a cars drivetrain. One line describes all of the points with a `4` value, one line describes all of the points with an `f` value, and one line describes all of the points with an `r` value.

We can make it more clear by overlaying the lines on top of the raw data and then coloring everything according to `drv`.

To display multiple geoms in the same plot, add multiple geom functions to `ggplot()`:
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

The package **ggplot2** provides over 30 geoms, and extension packages provide even more (see https://www.ggplot2-exts.org for a sampling). The best way to get a comprehensive overview is the **ggplot2** cheatsheet, which you can find at http://rstudio.com/cheatsheets. To learn more about any single geom, use help: `?geom_smooth`.

This introduces some duplication in our code. Imagine if we wanted to change the y-axis to display `cty` instead of `hwy`. We'd need to change the variable in two places, and we might forget to update one. We can avoid this type of repetition by passing a set of mappings to `ggplot()`, which will treat these mappings as global mappings that apply to each geom in the graph. In other words, this code will produce the same plot as the previous code:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

If you place mappings in a geom function, ggplot2 will treat them as local mappings for the layer. It will use these mappings to extend or overwrite the global mappings for that layer only. This makes it possible to display different aesthetics in different layers.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

We can use the same idea to specify different data for each layer. Here, our smooth line displays just a subset of the `mpg` data frame, the subcompact cars. The local data argument in `geom_smooth()` overrides the global data argument in `ggplot()` for that layer only.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```