---
title: "Making Graphs 2"
author: "James B. Elsner"
date: September 20, 2018
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

### Importing data

Consider the following data set showing all the olympic athletes during the modern era by event.
```{r}
ae.df <- read.csv(file = "http://myweb.fsu.edu/jelsner/temp/data/athlete_events.csv",
                  header = TRUE)
head(ae.df)
```

Each row is an athlete event. Athletes often compete in more than one event.

Suppose I asked you to plot the number of men and women athletes over time competing in the summer olympics. An athlete competing in more than one event gets counted only once. First we use some **dplyr** verbs.
```{r}
df <- ae.df %>%
  filter(Season == "Summer") %>%
  group_by(Year, Sex) %>%
  summarize(nA = n_distinct(Name))
```

Note: If we want to keep only rows (observations) that have a unique athlete then we would use `distinct(Name, .keep_all = TRUE)`

Next we make a plot.
```{r}
ggplot(data = df) +
  geom_point(mapping = aes(x = Year, y = nA, color = Sex)) +
  ylab("Number of Athletes Competing")
```

Fun examples. [Simpsons by the data](https://www.r-bloggers.com/the-simpsons-by-the-data/)

### Correlation

A scatter plot is a graphical depiction of correlation. Correlation describes the linear relationship between two variables with a single number.

Correlation is computed with the `cor()` function.
```{r}
cor(cars$speed, cars$dist)
```

The correlation computed on a set of data values a statistic. It describes whether larger- and smaller-than-average values of one variable are related to larger- or smaller-than-average values of the other variable.  If larger than average values of one variable tend to be associated with larger than average values of the other variable then the correlation is positive.

The scatter plot shows that the faster the car is moving the greater the stopping distance indicating a positive correlation.

A small arabic r is used to indicate correlation. The default correlation is called Pearson correlation and is defined as
$$
r = \frac{1}{n-1} \sum \left(\frac{x_i - \bar x}{s_x}\right)\left(\frac{y_i - \bar y}{s_y}\right)
$$
where $n$ is the sample size, $\bar x$ and $\bar y$ are the average of the x and y variables, respectively and $s_x$ and $s_y$ are the standard deviation of the x and y variables.

Correlation ranges between -1 and +1.

The correlation between x and y is the same as the correlation between y and x. We can check this by typing
```{r}
cor(cars$dist, cars$speed) == cor(cars$speed, cars$dist)
```

You can include the correlation on the scatter plot by adding a label layer (`geom_layer()` function). First assign the output from the correlation and `round()` functions. Then append this as a character string using the `paste()` function.
```{r}
r <- round(cor(cars$speed, cars$dist), 2)
p + geom_label(label = paste("Linear correlation is ", r), 
                   x = 10, y = 100)
```

The label is centered using the arguments `x =` and `y =` in the cartisian plane defined by the variables `speed` and `dist`. Note there is no 'hard' coding of the correlation value. Appearance is organic to the code sequence.

Consider the Old Faithful data frame (`faithful`). The correlation between wait time and eruption length is found by
```{r}
cor(faithful$eruptions, faithful$waiting)
```

Here interest centers on predicting eruption times based on waiting times. Are wait times correlated with eruption durations?
```{r}
ggplot(faithful, aes(x = waiting, y = eruptions)) + 
  geom_point() + 
  xlab("Wait Time (min)") + 
  ylab("Eruption Time (min)") +
  geom_density2d()
```

A 2-D kernel density layer is added to the plot with the `geom_density2d()` function.

The point density is shown with contours without units. You control the grid spacing and kernel width with the arguments `n =` and `h =`, respectively. If you want units, it is best to use the `kde2d()` function from the **MASS** package (part of the base install) and then use the `geom_contour()` function.

Is there a trend in hurricane landfalls?  One way to get an answer is to correlate the counts with the year.
```{r}
H <- read.table("http://myweb.fsu.edu/jelsner/temp/data/US.txt", 
                header = TRUE)
cor(H$Year, H$All)
```

Another example: The data set `kid.weights` (**UsingR**) contains height and weight data for children 0 to 12 years. `?kid.weights`.
```{r}
library(UsingR)
str(kid.weights)
ggplot(kid.weights, aes(x = height, y = weight)) + 
  geom_point()
```

As expected weight increases with height. However, the weight increase with height is more abrupt for the tallest kids suggesting a nonlinear relationship.  

In fact, body mass index (BMI) is a ratio of weight to squared height suggesting height and weight have a nonlinear relationship. Correlation only tells you about the linear relationship.

To see this add a linear and nonlinear trend line as layers on the plot with the `geom_smooth()` function.
```{r}
ggplot(kid.weights, aes(x = height, y = weight)) + 
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  geom_smooth(se = FALSE, color = "red")
```

The default is a local regression smoother, which creates the curved line. With the argument `method = lm` the blue (default color) line is the best fit straight line.

If the relationship between the variables is not linear it is often better to report the Spearman rank correlation. The rank correlation is the Pearson correlation applied to the ranks of the values.

That is, the data values are ordered from smallest to largest, and a particular data value's rank is its position after sorting, with 1 being the smallest and n being the largest, where n is the vector length. 
```{r}
x <- c(30, 20, 7, 42, 50, 21, 22)
rank(x)
```

The first three numbers are interpreted as: 30 is the fifth smallest value, 20 is the second and 7 is the smallest. If there are ties, the ranks are averaged.
```{r}
x <- c(30, 20, 7, 42, 50, 7, 22)
rank(x)
```

The rank correlation is obtained from the `cor()` function with the argument `method = "spearman"`.
```{r}
cor(kid.weights$weight, kid.weights$height)
cor(kid.weights$weight, kid.weights$height, method = "spearman")
```

As expected the the correlation is somewhat larger using ranks as the relationship appears to be more quadratic than linear.

Generate two vectors each with 10 random standard normal values. Verify that the rank correlation equals the Pearson correlation on the ranked values.
```{r}
x <- rnorm(10)
y <- rnorm(10)
cor(x, y, method = "spearman") == cor(rank(x), rank(y))
```

[Rank Correlation](http://en.wikipedia.org/wiki/Spearman's_rank_correlation_coefficient)

### Conditioning plots

Let's return to the kids weights data frame.
```{r}
head(kid.weights)
```

Note that by using the subset function `[]` you can list the weight for boys only
```{r}
kid.weights$weight[kid.weights$gender == "M"]
```

So the correlation between weight and height for the set of boys is
```{r}
cor(kid.weights$weight[kid.weights$gender == "M"], 
    kid.weights$height[kid.weights$gender == "M"])
```

Or using the syntax of the grammar for data.
```{r}
library(dplyr)
kid.weights %>%
  group_by(gender) %>%
  dplyr::summarize(r = cor(height, weight))
```

Let's look at the relationship between weight and height using a *conditioning* plot. Here we condition on the variable `gender`.
```{r}
ggplot(kid.weights, aes(x = height, y = weight)) + 
  geom_point() + 
  facet_grid(~ gender)
```

The conditioning is done using the `facet_grid()` function as a layer. Inside the function you specify the conditioning variable. The conditioning variable should be an integer, factor, or character. The tilde symbol (`~`) is read "depends on" or "conditioned on" or "function of". When the tilde is placed to the left of the conditioning variable then the plot is split left to right.

Another example: USA mortality rates (per one million inhabitants) for white males due to malignant melanoma 1950-1969.
```{r}
library(HSAUR2)
head(USmelanoma)
```

### Try it

1. Create a scatter plot showing male melanoma mortality on the y-axis (response variable) and latitude on the x-axis (explanatory variable).
2. Add a best-fit straight line on the plot. 
3. Condition the scatter plot on the variable 'ocean.' Is the relationship different for coastal states compared to non-coastal states?

```{r eval=FALSE}
ggplot(USmelanoma, aes(y = mortality, x = latitude )) + 
  geom_point() + 
  geom_smooth(method = lm) + 
  facet_grid(~ ocean )
```

Another example. Florida precipitation. Read the data then create a moltened data frame with the `melt()` function.
```{r}
L = "http://myweb.fsu.edu/jelsner/temp/data/FLprecip.txt"
FLp = read.table(L, na.string = "-9.900", header = TRUE)
head(FLp)
library(reshape2)
FLpL = melt(FLp, id.vars = "Year")
```

Time series by month.
```{r}
ggplot(FLpL, aes(x = Year, y = value)) + 
  geom_point() + 
  facet_wrap(~ variable, ncol = 12) + 
  geom_smooth(method = lm) +
  xlab("") + ylab("Precipitation (in)") +
  theme(axis.text.x = element_text(angle = 45))
```

### Graphs for communication

Adding labels turns our exploratory graphic into an expository graphic. For example consider the `mpg` dataset and plot highway mileage (`hwy`) as a function of engine size (`displ`) with the color of the point layer given by automobile class (`class`).
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
```

The purpose of a plot title is to summarise the main finding. Avoid titles that just describe what the plot is, e.g. "A scatterplot of engine displacement vs. fuel economy". If you need to add more text, there are two other useful labels.

* `subtitle =` adds additional detail in a smaller font beneath the title.
* `caption = ` adds text at the bottom right of the plot, often used to describe the source of the data.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov")
```

### Create a plot object

The assignment operator can be used to save the plot as an object. The object can be modified and printed.

Example: Given the 'cars' data frame from the **datasets** package. 
```{r}
str(cars)
```

Plot the car's stopping distance (`dist`) on the vertical (y) axis and the car's forward speed (`speed`) on the horizontal axis (x). Create the plot but assign it to an object called `p`.
```{r}
p <- ggplot(cars, aes(x = speed, y = dist)) +
        geom_point() +
        xlab("Forward Speed (mph)") +
        ylab("Stopping Distance (ft)")
p
p <- p + theme_bw()
p
p + geom_rug()
```

Notes:

1. the `aes()` function creates a list of variables to map. Here speed and dist from the data frame cars. The default argument list is ordered with the x variable first then the y variable, but it is good coding practice to name them using `x =` , etc.
2. The first layer is `geom_point()`. It maps the aesthetics to a cartesian plane. The `geom_point()` function needs an x and y aesthetic, but also understands `color`, `fill`, `shape`, `size`, `alpha`.
3. The second and third layers add labels to the axes. 
4. The graph is first saved as an object called `p` rather than rendered by the graphical device. It gets rendered by typing its name.
5. Additional layers are added to `p` using the `+` symbol.

The ggplot object `p` is a list of length 9 elements. Type
```{r, eval=FALSE}
str(p)
```

The first element is the data frame. The last element are the labels.

### Export a graph

When you knit to HTML and a plot is produced it gets output as a png file in a folder called "figure" in the directory where your .Rmd file is located.

You can also use the Export button under the Plots tab.

Or you can export the file directly using R code. Here the file gets put into your working directory.
```{r eval = FALSE, dev=c("pdf", "png")}
png(file = "MayNAO.png")
ggplot(NAO, aes(x = Year, y = May)) + 
  geom_line() + 
  xlab("Year") + 
  ylab("May NAO (s.d.)")
dev.off()
```

Note that the function `png()` opens the device and the function `dev.off()` closes it.  

Recall to list the files in your working directory type `dir()`.

My workflow. 

### Give it a try

Use the `cars` data frame (**datasets**) to create a histogram and box plot of stopping distance (column name `dist`). Also create a scatter plot of stopping distance (`dist`) versus `speed` with `speed` on the horizontal axis.

### Animation

Functions in the **gganimate** package add additional syntax to the grammar to allow simple animation.
```{r}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/FLprecip.txt"
FLp <- read.table(loc, 
                  na.string = "-9.900", 
                  header = TRUE)
```

```{r}
library(gganimate)
ggplot(FLp) +
  geom_point(mapping = aes(x = Year, y = Jan)) +
  ylab("January Precipitation in Florida (in)") +
  theme_minimal() +
  transition_time(Year)
```

```{r}
ggplot(FLp, mapping = aes(x = Year, y = Jan)) +
  geom_line() +
  geom_segment(mapping = aes(xend = 2011, yend = Jan), color = "blue") +
  geom_point() +
  geom_text(mapping = aes(x = 2020, label = Year)) +
  transition_reveal(Year, Year) +
  ease_aes('linear')
```

Note: here we put the mapping argument inside the `ggplot()` function. This allows all the layers to use these aesthetics. Layers that need additional aethetics get them through additional mapping arguments.

https://cran.r-project.org/web/packages/ggTimeSeries/vignettes/ggTimeSeries.html

## Problem Set #3