---
title: "Making Graphs 2"
author: "James B. Elsner"
date: September 20, 2018
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

"**...conditioning is not only something one does with ones hair, but something you also can do with data.**''---Rasmus Baath

After today you will know how to create graphs that display information from more than two variables.

### Facets

The scatter plot is the workhorse of statistical graphs. We can add a third variable by making the variable an aesthetic. We saw this last time.
```{r}
library(dplyr)
library(ggplot2)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

Another way to include a third variable, particularly useful for categorical variables, is to split the plot into facets, subplots that each display one subset of the data.

To facet a plot by a single variable, use `facet_wrap()`. The first argument of `facet_wrap()` should be a formula, which we create with `~` (tilde) followed by a variable name (here 'formula' is the name of a data structure in R, not a synonym for 'equation'). The variable that we pass to `facet_wrap()` should be discrete.

The tilde symbol (`~`) is read "depends on" or "conditioned on" or "function of". When the tilde is placed to the left of the conditioning variable then the plot is split left to right.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

To facet a plot on the combination of two variables, add `facet_grid()` to the plot call. The first argument of `facet_grid()` is also a formula. This time the formula should contain two variable names separated by a `~` with the first variable named varying in the vertical direction and the second varying in the horizontal direction. 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

Here `drv` refers to the drivetrain: front-wheel (`f`), rear-wheel (`r`) or 4-wheel (`4`).

To change the geom in our plot, change the geom function that we add to `ggplot()`. For instance,
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

Every geom function in **ggplot2** takes a mapping argument. However, not every aesthetic works with every geom. We could set the shape of a point, but we can't set the 'shape' of a line. On the other hand, we can set the line type of a line. `geom_smooth()` draws a different line, with a different linetype, for each unique value of the variable that we map to `linetype`.
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

Here `geom_smooth()` separates the cars into three lines based on their `drv` value, which describes a cars drivetrain. One line describes all of the points with a `4` value, one line describes all of the points with an `f` value, and one line describes all of the points with an `r` value.

We can make it more clear by overlaying the lines on top of the raw data and then coloring everything according to `drv`.

To display multiple geoms in the same plot, add multiple geom functions to `ggplot()`:
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

The package **ggplot2** provides over 30 geoms, and extension packages provide even more (see https://www.ggplot2-exts.org for a sampling). The best way to get a comprehensive overview is the **ggplot2** cheatsheet, which you can find at http://rstudio.com/cheatsheets. To learn more about any single geom, use help: `?geom_smooth`.

This introduces some duplication in our code. Imagine if we wanted to change the y-axis to display `cty` instead of `hwy`. We'd need to change the variable in two places, and we might forget to update one. We can avoid this type of repetition by passing a set of mappings to `ggplot()`, which will treat these mappings as global mappings that apply to each geom in the graph. In other words, this code will produce the same plot as the previous code:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

If you place mappings in a geom function, **ggplot2** will treat them as local mappings for the layer. It will use these mappings to extend or overwrite the global mappings for that layer only. This makes it possible to display different aesthetics in different layers.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

We can use the same idea to specify different data for each layer. Here, our smooth line displays just a subset of the `mpg` data frame, the subcompact cars. The local data argument in `geom_smooth()` overrides the global data argument in `ggplot()` for that layer only.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

### Another example

Consider the following data set showing all the olympic athletes during the modern era by event.
```{r}
ae.df <- read.csv(file = "http://myweb.fsu.edu/jelsner/temp/data/athlete_events.csv",
                  header = TRUE)
head(ae.df)
```

Each row is an athlete event. Athletes often compete in more than one event.

Suppose I asked you to plot the number of men and women athletes over time competing in the summer olympics. An athlete competing in more than one event gets counted only once. First we use some **dplyr** verbs.
```{r}
df <- ae.df %>%
  filter(Season == "Summer") %>%
  group_by(Year, Sex) %>%
  summarize(nA = n_distinct(Name))
```

Note: If we want to keep only rows (observations) that have a unique athlete then we would use `distinct(Name, .keep_all = TRUE)`

Next we make a plot.
```{r}
ggplot(data = df) +
  geom_point(mapping = aes(x = Year, y = nA, color = Sex)) +
  ylab("Number of Athletes Competing")
```

Fun examples. [Simpsons by the data](https://www.r-bloggers.com/the-simpsons-by-the-data/)

### Correlation

A scatter plot is a graphical depiction of correlation. Correlation describes the linear relationship between two variables with a single number.

Correlation is computed with the `cor()` function.
```{r}
cor(cars$speed, cars$dist)
```

The correlation computed on a set of data values a statistic. It describes whether larger- and smaller-than-average values of one variable are related to larger- or smaller-than-average values of the other variable.  If larger than average values of one variable tend to be associated with larger than average values of the other variable then the correlation is positive.

The scatter plot shows that the faster the car is moving the greater the stopping distance indicating a positive correlation.

A small arabic r is used to indicate correlation. The default correlation is called Pearson correlation and is defined as
$$
r = \frac{1}{n-1} \sum \left(\frac{x_i - \bar x}{s_x}\right)\left(\frac{y_i - \bar y}{s_y}\right)
$$
where $n$ is the sample size, $\bar x$ and $\bar y$ are the average of the x and y variables, respectively and $s_x$ and $s_y$ are the standard deviation of the x and y variables.

Correlation ranges between -1 and +1.

The correlation between x and y is the same as the correlation between y and x. We can check this by typing
```{r}
cor(cars$dist, cars$speed) == cor(cars$speed, cars$dist)
```

We can include the correlation on the scatter plot by adding a label layer (`geom_layer()` function). First assign the output from the correlation and `round()` functions. Then append this as a character string using the `paste()` function.
```{r}
r <- round(cor(cars$speed, cars$dist), 2)
p + geom_label(label = paste("Linear correlation is ", r), 
                   x = 10, y = 100)
```

The label is centered using the arguments `x =` and `y =` in the cartisian plane defined by the variables `speed` and `dist`. Note there is no 'hard' coding of the correlation value. Appearance is organic to the code sequence.

Consider the Old Faithful data frame (`faithful`). The correlation between wait time and eruption length is found by
```{r}
cor(faithful$eruptions, faithful$waiting)
```

Here interest centers on predicting eruption times based on waiting times. Are wait times correlated with eruption durations?
```{r}
ggplot(faithful, aes(x = waiting, y = eruptions)) + 
  geom_point() + 
  xlab("Wait Time (min)") + 
  ylab("Eruption Time (min)") +
  geom_density2d()
```

A 2-D kernel density layer is added to the plot with the `geom_density2d()` function.

The point density is shown with contours without units. You control the grid spacing and kernel width with the arguments `n =` and `h =`, respectively. If you want units, it is best to use the `kde2d()` function from the **MASS** package (part of the base install) and then use the `geom_contour()` function.

Is there a trend in hurricane landfalls?  One way to get an answer is to correlate the counts with the year.
```{r}
H <- read.table("http://myweb.fsu.edu/jelsner/temp/data/US.txt", 
                header = TRUE)
cor(H$Year, H$All)
```

Another example: The data set `kid.weights` (**UsingR**) contains height and weight data for children 0 to 12 years. `?kid.weights`.
```{r}
library(UsingR)
str(kid.weights)
ggplot(kid.weights, aes(x = height, y = weight)) + 
  geom_point()
```

As expected weight increases with height. However, the weight increase with height is more abrupt for the tallest kids suggesting a nonlinear relationship.  

In fact, body mass index (BMI) is a ratio of weight to squared height suggesting height and weight have a nonlinear relationship. Correlation only tells you about the linear relationship.

To see this add a linear and nonlinear trend line as layers on the plot with the `geom_smooth()` function.
```{r}
ggplot(kid.weights, aes(x = height, y = weight)) + 
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  geom_smooth(se = FALSE, color = "red")
```

The default is a local regression smoother, which creates the curved line. With the argument `method = lm` the blue (default color) line is the best fit straight line.

If the relationship between the variables is not linear it is often better to report the Spearman rank correlation. The rank correlation is the Pearson correlation applied to the ranks of the values.

That is, the data values are ordered from smallest to largest, and a particular data value's rank is its position after sorting, with 1 being the smallest and n being the largest, where n is the vector length. 
```{r}
x <- c(30, 20, 7, 42, 50, 21, 22)
rank(x)
```

The first three numbers are interpreted as: 30 is the fifth smallest value, 20 is the second and 7 is the smallest. If there are ties, the ranks are averaged.
```{r}
x <- c(30, 20, 7, 42, 50, 7, 22)
rank(x)
```

The rank correlation is obtained from the `cor()` function with the argument `method = "spearman"`.
```{r}
cor(kid.weights$weight, kid.weights$height)
cor(kid.weights$weight, kid.weights$height, method = "spearman")
```

As expected the the correlation is somewhat larger using ranks as the relationship appears to be more quadratic than linear.

Generate two vectors each with 10 random standard normal values. Verify that the rank correlation equals the Pearson correlation on the ranked values.
```{r}
x <- rnorm(10)
y <- rnorm(10)
cor(x, y, method = "spearman") == cor(rank(x), rank(y))
```

[Rank Correlation](http://en.wikipedia.org/wiki/Spearman's_rank_correlation_coefficient)


### Another example

Let's look at the kids weights data frame from the **UsingR** package.
```{r}
library(UsingR)
head(kid.weights)
```

Note that by using the subset function `[]` you can list the weight for boys only
```{r}
kid.weights$weight[kid.weights$gender == "M"]
```

So the correlation between weight and height for the set of boys is
```{r}
cor(kid.weights$weight[kid.weights$gender == "M"], 
    kid.weights$height[kid.weights$gender == "M"])
```

Or using the syntax of the grammar for data.
```{r}
library(dplyr)
kid.weights %>%
  group_by(gender) %>%
  dplyr::summarize(r = cor(height, weight))
```

Let's look at the relationship between weight and height using a facet plot (conditioning) plot. Here we facet (condition) on the variable `gender`.
```{r}
ggplot(kid.weights, aes(x = height, y = weight)) + 
  geom_point() + 
  facet_grid(~ gender)
```

Another example: USA mortality rates (per one million inhabitants) for white males due to malignant melanoma 1950-1969.
```{r}
library(HSAUR2)
head(USmelanoma)
```

Workflow

1. Create a scatter plot showing male melanoma mortality on the y-axis (response variable) and latitude on the x-axis (explanatory variable).
2. Add a best-fit straight line on the plot. 
3. Condition the scatter plot on the variable 'ocean.' Is the relationship different for coastal states compared to non-coastal states?

```{r eval=FALSE}
ggplot(USmelanoma, aes(y = mortality, x = latitude )) + 
  geom_point() + 
  geom_smooth(method = lm) + 
  facet_grid(~ ocean )
```

Another example. Florida precipitation. Read the data then create a moltened data frame with the `melt()` function.
```{r}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/FLprecip.txt"
FLp <- read.table(L, 
                  na.string = "-9.900", 
                  header = TRUE)
head(FLp)
library(reshape2)
FLpL <- melt(FLp, id.vars = "Year")
```

Time series by month.
```{r}
ggplot(FLpL, aes(x = Year, y = value)) + 
  geom_point() + 
  facet_wrap(~ variable, ncol = 12) + 
  geom_smooth(method = lm) +
  xlab("") + ylab("Precipitation (in)") +
  theme(axis.text.x = element_text(angle = 45))
```

### Graphs for communication

Adding labels turns our exploratory graphic into an expository graphic. For example consider the `mpg` dataset and plot highway mileage (`hwy`) as a function of engine size (`displ`) with the color of the point layer given by automobile class (`class`).
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
```

The purpose of a plot title is to summarise the main finding. Avoid titles that just describe what the plot is, e.g. "A scatterplot of engine displacement vs. fuel economy". If you need to add more text, there are two other useful labels.

* `subtitle =` adds additional detail in a smaller font beneath the title.
* `caption = ` adds text at the bottom right of the plot, often used to describe the source of the data.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov")
```

### Create a plot object

The assignment operator can be used to save the plot as an object. The object can be modified and printed.

Example: Given the 'cars' data frame from the **datasets** package. 
```{r}
str(cars)
```

Plot the car's stopping distance (`dist`) on the vertical (y) axis and the car's forward speed (`speed`) on the horizontal axis (x). Create the plot but assign it to an object called `p`.
```{r}
p <- ggplot(cars, aes(x = speed, y = dist)) +
        geom_point() +
        xlab("Forward Speed (mph)") +
        ylab("Stopping Distance (ft)")
p
p <- p + theme_bw()
p
p + geom_rug()
```

Notes:

1. the `aes()` function creates a list of variables to map. Here speed and dist from the data frame cars. The default argument list is ordered with the x variable first then the y variable, but it is good coding practice to name them using `x =` , etc.
2. The first layer is `geom_point()`. It maps the aesthetics to a cartesian plane. The `geom_point()` function needs an x and y aesthetic, but also understands `color`, `fill`, `shape`, `size`, `alpha`.
3. The second and third layers add labels to the axes. 
4. The graph is first saved as an object called `p` rather than rendered by the graphical device. It gets rendered by typing its name.
5. Additional layers are added to `p` using the `+` symbol.

The ggplot object `p` is a list of length 9 elements. Type
```{r, eval=FALSE}
str(p)
```

The first element is the data frame. The last element are the labels.

### Export a graph

When we knit to HTML and a plot is produced it gets output as a png file in a folder called "figure" in our working directory.

We can use the Export button under the Plots tab.

Or we can export the file directly using R code. Here the file gets put into our working directory.
```{r eval = FALSE, dev=c("pdf", "png")}
png(file = "MayNAO.png")
ggplot(NAO, aes(x = Year, y = May)) + 
  geom_line() + 
  xlab("Year") + 
  ylab("May NAO (s.d.)")
dev.off()
```

Note that the function `png()` opens the device and the function `dev.off()` closes it.  

Recall: list the files in your working directory with the command `dir()`.

## Problem Set #3