---
title: "Maps in R"
author: "James B. Elsner"
date: September 25, 2018
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

### Animation

Functions in the **gganimate** package add additional syntax to allow simple animation.
```{r}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/FLprecip.txt"
FLp <- read.table(loc, 
                  na.string = "-9.900", 
                  header = TRUE)
```

```{r}
#devtools::install_github('thomasp85/gganimate')
library(gganimate)
ggplot(FLp) +
  geom_point(mapping = aes(x = Year, y = Jan)) +
  ylab("January Precipitation in Florida (in)") +
  theme_minimal() +
  transition_time(Year)
```

```{r}
ggplot(FLp, mapping = aes(x = Year, y = Jan)) +
  geom_line() +
  geom_segment(mapping = aes(xend = 2011, yend = Jan), color = "blue") +
  geom_point() +
  geom_text(mapping = aes(x = 2020, label = Year)) +
  transition_reveal(Year, Year) +
  ease_aes('linear')
```

Note: here we put the mapping argument inside the `ggplot()` function. This allows all the layers to use these aesthetics. Layers that need additional aethetics get them through additional mapping arguments.

### Time series plots

https://cran.r-project.org/web/packages/ggTimeSeries/vignettes/ggTimeSeries.html

## Maps with ggplot2

R has a relatively simple mechanism, through the **maps** package, for making outlines of maps and plotting lat-long points and paths on them.

With the advent of packages like **sp**, **rgdal**, **rgeos**, and **sf**. R has acquired some of the functionality of traditional GIS packages.

Load the libraries we will use
```{r}
library(maps)
library(mapdata)
```

The **maps** package contains outlines of continents, countries, states, and counties that have been with R for a long time. It comes with a plotting function but we will use **ggplot2**. The **mapdata** package contains a few more, higher-resolution outlines.

Since the **ggplot2** functions work on data frames we first translate the **maps** data into data frames with the `map_data()` function. The syntax is `map_data("name")` where `"name"` is a quoted string of the name of a map in the **maps** or **mapdata** package.
```{r}
usa <- map_data("usa")
dim(usa)
```

Or a  high-res world map centered on the Pacific Ocean from mapdata
```{r}
w2hr <- map_data("world2Hires")
dim(w2hr)
```

Structure of the data frame.
```{r}
str(usa)
```

* `long` is longitude. Boundary points to the west of the prime meridian are negative.
* `lat` is latitude.
* `order` is the order `ggplot()` uses to connect the points
* `region` and `subregion` are the country and larger region
* `group` is very important! **ggplot2** functions can take a group argument which controls whether adjacent points should be connected by lines. If they are in the same group they get connected, otherwise not.

Maps are plotted with the `geom_polygon()` which draws lines between the boundary points and closes them up by drawing a line from the last point to the first point.

The `group =` aesthetic must include the `group` variable from the data frame. 
```{r}
ggplot(data = usa) +
  geom_polygon(aes(x = long, 
                   y = lat, 
                   group = group))
```

By default, `geom_polygon()` draws with no line color and with a black fill.

Note that the range of longitudes is larger than the range of latitudes. We change this with the function `coord_fixed()` which fixes the relationship between one unit in the y direction and one unit in the x direction. This keeps the aspect ratio of the plot the same.

Make another map. This time get rid of the fill and use a gray border.
```{r}
ggplot(data = usa) + 
  geom_polygon(aes(x = long, 
                   y = lat, 
                   group = group), 
               fill = NA, 
               color = "gray") + 
  coord_fixed(1.3)
```


### Adding locations.

Let's add the location of Tallahassee and Milwaukee.
```{r}
Cities <- data.frame(long = c(-84.2809, -87.9735),
                     lat = c(30.4381,43.0115),
                     names = c("Tallahassee", "Milwaukee"))
ggplot(data = usa) + 
  geom_polygon(aes(x = long, 
                   y = lat,  
                   group = group), 
               fill = NA, 
               color = "gray") + 
  coord_fixed(1.3) + 
  geom_point(data = Cities, aes(x = long, y = lat), color = "black", size = 3)
```

### State maps

We can also get a data frame of polygons for each state.
```{r}
states <- map_data("state")
dim(states)
```

```{r}
ggplot(data = states) + 
  geom_polygon(aes(x = long, 
                   y = lat, 
                   fill = region, 
                   group = group), 
               color = "gray") + 
  coord_fixed(1.3) +
  guides(fill = FALSE)  # do this to leave off the color legend
```

```{r}
library(dplyr)
southeast <- states %>%
  filter(region == "florida" | region == "georgia" | region == "alabama")

ggplot(data = southeast) + 
  geom_polygon(aes(x = long, 
                   y = lat, 
                   group = group), fill = "green", 
               color = "gray") + 
  coord_fixed(1.1)
```

### County boundaries

Let's zoom in on Florida and look at counties. First lets get the state boundary.
```{r}
florida_st <- states %>%
  filter(region == "florida")
```

Next let's get the county boundaries.
```{r}
florida_ct <- map_data("county") %>%
  filter(region == "florida")
head(florida_ct)
```

Let's make a base map.
```{r}
FL_base <- ggplot(data = florida_st, 
                  mapping = aes(x = long, 
                                y = lat, 
                                group = group)) + 
  coord_fixed(1.1) + 
  geom_polygon(color = "black", fill = "gray")
FL_base
```

We remove the grid lines using the `theme_void()`.
```{r}
FL_base +
  theme_void()
```

Now add the county boundaries in white. We add the state border back on top to overwrite the white borders from the county data frame.
```{r}
FL_base + 
  theme_void() + 
  geom_polygon(data = florida_ct, 
               fill = NA, 
               color = "white") +
  geom_polygon(color = "black",
               fill = NA) 
```

### Thematic maps

We make a thematic (choropleth) map by adding a variable name to the `fill =` within the asthetic function.

First let's look at the subregion (county) names.
```{r}
florida_ct %>%
  distinct(subregion)
```

The names need to match in the file containing the variable we want to choropleth. I've done this and put it into a text file.
```{r}
FLpop <- read.table(file = "http://myweb.fsu.edu/jelsner/temp/data/FL_ct_pop2017.txt",
                    header = TRUE,
                    stringsAsFactors = FALSE)
```

Next we join the population list to the boundary points data.
```{r}
florida_ctJ <- left_join(florida_ct, 
                         FLpop, 
                         by = "subregion")
```

Then we make the map. We reuse the code from above replacing the `fill = NA` with `aes(fill = population)`. `population` is the name of the variable in the joined data frame that we want to map. Since the range of population values is large we scale the fill by the logarithm base 10.
```{r}
FL_base +  
  theme_void() + 
  geom_polygon(data = florida_ctJ, 
               aes(fill = population), 
               color = "gray") +
  geom_polygon(color = "black",
               fill = NA) +
  scale_fill_gradient(trans = "log10", 
                      name = "Population (2017)",
                      label = scales::comma)
```

## US maps including Alaska and Hawaii

https://cran.r-project.org/web/packages/usmap/vignettes/introduction.html

Plots of US maps in R usually lack Alaska and Hawaii. Plotting takes the literal longitude and latitude coordinates and maps it to a cartesian x-y coordinate graph. 

Alaska and Hawaii are very far from the mainland US when using this so it can be unwieldy to include them. The **usmap** package solves this issue by providing data frames which have Alaska and Hawaii moved to a convenient spot just to the bottom left of the contiguous United States.

Blank US state map.
```{r}
library(usmap)
plot_usmap()
```
With county borders.
```{r}
plot_usmap(regions = "counties")
```

The raw US map boundary points for counties or states can be obtained for further manipulation (and joining with data). The default regions is "states".
```{r}
states_df <- usmap::us_map()
counties_df <- usmap::us_map(regions = "counties")
head(states_df)
```

### FIPS codes

FIPS codes are defined in the Federal Information Processing Standards by the US government. One usage is uniquely identifying US states and counties (among other things such as identifying countries for the CIA World Factbook). Downloading datasets from the US Census will often include FIPS codes as identifiers so it can be helpful to know what a FIPS code represents. 

Functions in **usmap** are built around the FIPS code identification system and so convenience methods for accessing them and performing reverse-lookups have been included.

State/County FIPS lookup
```{r}
fips(state = "FL")
```
```{r}
fips(state = "Florida")
```

Get FIPS code for a county.
```{r}
fips(state = "FL", 
     county = "Leon")
```

The argument values are NOT case sensitive!
```{r}
fips(state = "fl", 
     county = "leON")
```

FIPS reverse lookup. If the FIPS code is known and want to see what state/county it corresponds to, use the reverse lookup function fips_info.
```{r}
fips_info(c("12", "33", "34"))
```

Or
```{r}
fips_info(c("01001", "01003", "01005", "01007"))
```

The nice thing about `plot_usmap()` is it returns a **ggplot2** object, which means we can add layers to the plot as we do with graphs.

```{r}
plot_usmap(regions = "counties") + 
  labs(title = "US Counties", 
       subtitle = "This is a blank map of the counties of the United States.") + 
  theme(panel.background = element_rect(colour = "black", fill = "lightblue"))
```

```{r}
plot_usmap(include = c("FL", "GA", "AL", "MS", "SC", "TN")) +
  labs(title = "Southeaster US States", subtitle = "These are the states in the Southeast.")
```

### Add some data

The **usmap** comes with some data files including population and percent poverty by state and county. For example,
```{r}
data(package = "usmap")
head(statepop)
```

```{r}
plot_usmap(data = statepop, 
           values = "pop_2015", 
           lines = "gray") + 
  scale_fill_continuous(name = "Population (2015)", label = scales::comma) + 
  theme(legend.position = "right")
```

## Maps with tmap

https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html

The **tmap** package is a flexible, layer-based, and easy-to-use approach to create thematic maps including choropleths and bubble maps. It is based on the grammar of graphics, and resembles the syntax of **ggplot2**.
With the **tmap** package, thematic maps can be generated with greater flexibility.

```{r}
library(tmap)
data("World")
tm_shape(World) +
    tm_polygons("HPI")
```

`World` is a spatial object of class `sf` from the **sf** package; it is a data frame with a special column that contains a geometry for each row, in this case polygons. 
```{r}
class(World)
```

```{r}
head(World)
```

In order to plot the geometry in **tmap**, we need to specify it with `tm_shape()` function. Plotting layers are then added with the `+` operator, in this case `tm_polygons()`. 

There are many layer functions in **tmap**, which can be found in the documentation by their `tm_` prefix. See also `?'tmap-element'`.

### Simple feature data frames

Here we get the tornado data directly from the Storm Prediction Center (SPC) http://www.spc.noaa.gov/gis/svrgis/ as a shapefile.
```{r}
download.file("http://www.spc.noaa.gov/gis/svrgis/zipped/tornado.zip",
              "tornado.zip", mode = "wb")
unzip("tornado.zip")
```

The data are downloaded as a zipped file then unzipped using the `unzip()` function. This results in a folder called `torn` containing the shapefiles `torn.*`

Shapefiles can be imported in several ways. Here we use the `st_read()` function from the **sf** package.
```{r}
library(sf)
sfdf <- st_read(dsn = "torn", 
               layer = "torn", 
               stringsAsFactors = FALSE)
```

The result is a simple feature data frame with 62520 features (observations) and 21 fields (variables). Each observation is a unique tornado. Let's first glimpse the data frame. Here we use the `glimpse()` function from the **dplyr** package.
```{r}
glimpse(sfdf)
```

The first 21 columns are variables (attributes). The last column contains the geometry. The `geometry` column is in well known text (WKT) format. Each tornado is a `LINESTRING` with a start and end location. This is where the `tm_shape()` function will look for the mapping information.

For example, let's map the tracks of all tornadoes since 2011.
```{r}
sfdf <- sfdf %>%
  filter(yr > 2010) 

tm_shape(World) +
  tm_polygons() +
tm_shape(sfdf) +
    tm_lines(col = "blue")
```

The first `tm_shape()` defines the base layer which defines the map extent. To change that we can use the `is.master = TRUE` argument in the second `tm_shape()` function.
```{r}
tm_shape(World) +
  tm_polygons() +
tm_shape(sfdf, is.master = TRUE) +
    tm_lines(col = "blue")
```

## Thematic maps 

Let's try something a bit more complicated. Suppose we want the number of tornadoes originating in each state on a choropleth map. We first group and summarize.
```{r}
sfdf2 <- sfdf %>%
  filter(st != "PR") %>%
  group_by(st) %>%
  summarize(nT = n())
glimpse(sfdf2)
```

The resulting data frame contains the grouped-by column `st` (origin state) and the corresponding number of tornadoes. There were 415 tornadoes in Alabama since 2011.

Next we get a simple feature containing the geometry of the U.S. states with Alaska and Hawaii inserted. We get this from Bob Rudis's  **albersusa** package available on GitHub. Make sure you have the **devtools** package installed.
```{r}
#devtools::install_github("hrbrmstr/albersusa")
library(albersusa)
us_sf <- usa_sf("aeqd") 
glimpse(us_sf)
```

Each feature is a state (or district). The attributes (variables) include the state name, abbreviation, and population estimates. The geometry is MULTIPOLYGON. This is needed because of islands, etc.
```{r}
tm_shape(us_sf) +
  tm_polygons() 
```

Next we need to change the state name from `iso_3166_2` to `st`.
```{r}
us_sf <- us_sf %>%
 mutate(st = as.character(iso_3166_2))
```

Then we join the tornado counts with the map simple feature object. The `left_join()` function only works on data frames (not simple feature data frames) so we first convert them to data frames. We then convert the resulting data frame back to a simple feature data frame with the `st_as_sf()` function. Finally we select only the number of tornadoes column (`nT`).
```{r}
sfdf3 <- left_join(as.data.frame(us_sf), 
                   as.data.frame(sfdf2), 
                   by = "st") %>%
         st_as_sf() %>%
         select(nT)
head(sfdf3)
```

NOTE: we want the result of the join to be the MULTIPOLGON geometry associated with each state so the first argument is the `us_sf` simple feature data frame.

Next we use the functions in the **tmap** package. 
```{r}
tm_shape(sfdf3) +
  tm_polygons("nT", 
           title = "Tornado Counts",
           palette = "Reds")
```

We can improve the defaults with additional layers including text, compass, and scale bar. The last layer is the print view.
```{r}
tm_shape(sfdf3) +
  tm_polygons("nT", 
              border.col = NULL,
              title = "Tornado Counts",
              palette = "Reds") +
  tm_text("nT", size = .5) +
  tm_compass() + 
  tm_scale_bar(lwd = .5)
```

The format of the **tmap** map objects (meoms) are like those of the **ggplot2** geometric objects (geoms) making it easy to get to a publication-quality map. The fine details can be worked out in production.

More information? See: https://cran.r-project.org/web/packages/tmap/vignettes/tmap-nutshell.html