---
title: "Exam 1 Review"
author: "James B. Elsner"
date: October 2, 2018
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

**1** Use base R functions to: (a) Read the tornado dataset (http://myweb.fsu.edu/jelsner/temp/data/Tornadoes.txt) as a spreadsheet into your R session. (b) How many rows and how many columns are in the data frame? (c) List the column names.
```{r}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/Tornadoes.txt"
Torn.df <- read.table(loc, header = TRUE)
dim(Torn.df)
names(Torn.df)
```

**2** Use base R functions to: (a) Create a vector called `inj` by assigning to it the `INJURIES` column from the tornado dataset, then answer the following questions: (b) What is the average per tornado number of injuries? (c) What is the median per tornado number of injuries? (d) What is the maximum number of injuries? (e) What are the 95th and 99th percentile number of injuries? (f) How many tornadoes had two or more injuries? (g) What percentage of the total number of tornadoes is that?
```{r}
inj <- Torn.df$INJURIES
mean(inj)
median(inj)
max(inj)
quantile(inj, prob = c(.95, .99))
sum(inj >= 2)
sum(inj >= 2)/length(inj) * 100
```

**3** Use the functions from the **dplyr** package to answer the following: (a) Determine the number of tornadoes by month. (b) What state has had the most tornadoes during the month of March? (c) What state has had the most tornadoes with damage ratings (`FSCALE`) exceeding 3? (d) What state has had the most fatalities (`FATALITIES`)?
```{r}
library(dplyr)
Torn.df %>%
  group_by(MONTH) %>%
  summarize(nT = n())
```

```{r}
Torn.df %>%
  filter(MONTH == 3) %>%
  group_by(STATE) %>%
  summarize(nT = n()) %>%
  arrange(desc(nT))
```

```{r}
Torn.df %>%
  filter(FSCALE > 3) %>%
  group_by(STATE) %>%
  summarize(nT = n()) %>%
  arrange(desc(nT))
```

```{r}
Torn.df %>%
  group_by(STATE) %>%
  summarize(nF = sum(FATALITIES)) %>%
  arrange(desc(nF))
```

**4** Use the functions from the **dplyr** and **ggplot2** packages to make a time-series graph of annual average number of fatalities. Include a trend line.
```{r}
library(ggplot2)
Torn.df %>%
  group_by(YEAR) %>%
  summarize(nF = sum(FATALITIES)) %>%
  ggplot(., aes(x = YEAR, y = nF)) +
    geom_point() +
    geom_smooth(method = lm)
```

**5** Use the `cars` data frame (**datasets**) to create a histogram and box plot of stopping distance (column name `dist`). Also create a scatter plot of stopping distance (`dist`) versus `speed` with `speed` on the horizontal axis.
